Glia-mediated gut-brain cytokine signaling couples sleep to intestinal inflammation

Version 1

Abstract

Abstract

Sickness-induced sleep is a behavior conserved across species that promotes recovery from illness, yet the underlying mechanisms are poorly understood. Here, we show that interleukin-6-like cytokine signaling from the Drosophila gut to brain glial cells regulates sleep. Under healthy conditions, this pathway promotes wakefulness. However, elevated gut cytokine signaling in response to oxidative stress – triggered by immune and inflammatory responses in the intestine – induces sleep. The cytokines Unpaired 2 and -3 are upregulated by oxidative stress in enteroendocrine cells and activate JAK-STAT signaling in glial cells, including those of the blood-brain barrier (BBB). This activity maintains elevated sleep during oxidative-stress-induced intestinal disturbances, suggesting that the JAK-STAT pathway in glia inhibits wake-promoting signaling to facilitate sleep-dependent restoration under these conditions. We find that the enteric peptide Allatostatin A (AstA) enhances wakefulness, and during intestinal oxidative stress, gut-derived Unpaired 2/3 inhibits AstA receptor expression in BBB glia, thereby sustaining an elevated sleep state during gut inflammation or illness. Taken together, our work identifies a gut-to-glial communication pathway that couples sleep with intestinal homeostasis and disease, enhancing sleep during intestinal sickness, and contributes to our understanding of how sleep disturbances arise from gastrointestinal disturbances.

Introduction

Sleep is a conserved behavior essential for physical health and mental well-being. This process maintains physiological balance and promotes recovery from illnesses and other stressors1,2. In healthy states, animals exhibit rhythmic periods of wakefulness and activity, but their sleep patterns significantly increase during illness. This adaptive behavior is related to the fundamental role of sleep in the recovery process, allowing the body to conserve energy and allocate resources towards eliminating pathogens and repairing tissue damage. Sleep patterns are generated by neural circuits within the brain. These circuits engage in complex interactions involving diverse brain regions, neurotransmitters, and signaling pathways to regulate the cycles of sleep and wakefulness3,4. For sleep to be effectively modulated during illness, there must be a dynamic interaction between the physiological states of the body’s organs and these central sleep-regulatory circuits. However, the signals that mediate this communication and the mechanisms by which they modulate sleep during health and disease remain poorly defined.

Sickness-induced sleep, a behavior conserved across species including mammals and flies5,6, is influenced by cytokines, which are key mediators of immune and inflammatory responses1,2. Cytokines such as Interleukin 1 (IL-1) and tumor necrosis factor alpha (TNFα) are expressed in the healthy mammalian brain in regions that are implicated in sleep regulation, and their circulating levels change during the normal sleep-wake cycle, peaking during the sleep phase. Furthermore, these factors’ effects on sleep appear to be dose-dependent, as low levels of IL-1 can enhance sleep, whereas higher doses can inhibit sleep, indicating a dual functionality. Since immune responses alter the expression of these cytokines, they have been hypothesized to act as “somnogens” that promote sleep during times of infection or illness. However, the connection between sleep and immune function is bidirectional, since sleep deprivation in mammals has been linked to increased inflammatory response via IL-1 and TNFα2. In the fruit fly Drosophila, sleep deprivation also seems to influence TNFα in astrocyte-like cells to regulate homeostatic sleep responses that enable sleep rebound after deprivation7. However, the effect of these cytokines on sleep has mostly been linked to their central expression and function within the central nervous system (CNS), while the coupling of cytokines produced by peripheral tissues to sleep-regulatory systems within the brain remains poorly understood.

Disorders affecting the gastrointestinal tract can lead to sleep disturbances8, which are also associated with virtually all mental illnesses9–11. Conditions including depression, anxiety, and disturbed sleep are frequently observed in individuals with gut inflammation, and the gut microbiome has also been linked to sleep quality and mental health8,12–14. These associations suggest a strong connection between gut health and sleep. For changes in gut status to bring about behavioral changes, the gut must sense its state of health, damage, or presence of pathogens and release signals that lead to altered cellular function within the brain. This gut-to-brain signaling is mediated in large part by hormonal factors released from specialized endocrine cells of the gut, the enteroendocrine cells (EECs)15,16. Like the mammalian intestine, the Drosophila gut produces numerous diverse hormones from specialized EECs17–22. Some of these gut hormones are released in response to nutritional intake, and they diet-dependently modulate sleep patterns and arousability through communication with neuroendocrine centers and brain circuits23–25.

In the fly, enteric infection or damage leads to the production of reactive oxygen species (ROS) and the increased expression of the IL-6-like inflammatory cytokines Unpaired 2 and -3 (Upd2/3) in the absorptive enterocytes, a response required for local gut regeneration26,27. These cytokine factors signal through their receptor Domeless (Dome) to activate the JAK/STAT signaling pathway, which is important for both immune function and metabolism in flies, demonstrating a conserved function of cytokine action in this species. While the three related cytokines, Upd1, Upd2, and Upd3, all signal through Dome, Upd2 and in particular Upd3 are IL-6-like cytokines mainly triggered by infection and are directly linked with cellular immune responses28–30. As in mammals, cytokines are also produced centrally within the Drosophila brain, and neuronal Upd1 acts in a leptin-like manner to regulate feeding31, a behavior that is also linked with sleep4. Peripheral cytokine signaling has also been shown to modulate sleep in this species, where the fat tissue releases Upd2 to reflect adequate nutrition, and this signal modulates sleep32. Furthermore, Unpaired cytokines have been implicated in the modulation of feeding behavior through effects on glial cells33. Glial cells, including those making up the blood-brain barrier (BBB), have recently gained attention for their role in sleep regulation in both flies and mammals34–37. Neurons in the CNS are separated from the circulatory system by the BBB, a selectively semi-permeable cell layer38, which presents a challenge for peripheral hormones to enter and signal to neurons within the brain. However, glial cells within the BBB are ideally positioned to receive and integrate systemic signals from peripheral organs and modulate neuronal function, thereby relaying peripheral information into the brain.

Here, we demonstrate Upd2 and Upd3 cytokine signaling from endocrine EECs in the intestine in Drosophila. Our findings show that Upd2/3 signaling from the EECs to BBB glial cells, plays a dual role in sleep regulation. Under normal, healthy conditions, EEC-derived Unpaired signaling sustains wakefulness, whereas in response to oxidative stress that leads to gut inflammation, elevated Unpaired signaling instead promotes sleep. Stress-induced EEC-derived Upd2/3 activates the JAK-STAT pathway in glial cells at the blood-brain interface and adjusts sleep through this activation based on intestinal homeostasis and levels of inflammatory signaling from the gut. Our results suggest that gut-derived Unpaired signaling influences sleep regulation through glial gating of wake-promoting AstA-mediated signals, thus linking intestinal health with CNS-dependent behaviors. These results identify a gut-brain connection by which gut disease impacts sleep regulation.

Results

Gut-derived Unpaired cytokine signaling regulates sleep

To investigate whether cytokine signaling from the gut regulates sleep, we silenced the expression of upd2 and upd3 in the EECs, which are a principal endocrine cell type in the gut that releases signals with systemic effects and constitutes the functional basis of gut-brain signaling. Using voilà-GAL4 (a driver that targets all EECs) to drive RNAi in EECs in conjunction with Tub-GAL80ts (hereafter referred to as voilà>) for temperature-induced RNAi induction exclusively in the adult stage to prevent developmental effects, we observed significant knockdown of the main IL-6 cytokine, upd330, in dissected adult female midguts (Fig. 1a), demonstrating that upd3 is expressed in EECs under normal homeostatic conditions. To eliminate potential neuron-derived phenotypes, we employed R57C10-GAL80, a form of nSyb-GAL80 that effectively inhibits neuronal GAL4 activity, in combination with Tub-GAL80ts and voilà-GAL425,39. To evaluate the effectiveness of this temperature-sensitive EEC-specific driver (referred to as EEC> hereafter) upon adult-restricted induction, we examined midgut upd2 and upd3 transcript levels and observed significant knockdown of both cytokines, which was reproduced with a second independent RNAi line targeting upd3 (Fig. 1b). Additionally, GFP expressed under the control of an upd3-GAL4 driver containing upd3 enhancer sequences, was apparent in midgut EECs, marked by staining against the EEC fate determinant Prospero (Fig. 1c). These results show that EECs of the adult female midgut are a source of Upd2 and Upd3 under normal conditions.

EEC-derived Unpaired signaling regulates sleep.

a, upd3 expression levels in midguts RNAi-mediated knockdown of upd3 in EECs using voilá-GAL4 in combination with Tubulin-GAL80ts (voilá>) (N=7). b, upd2 and upd3 expression levels in midguts in animals with EEC knockdown of upd2 or upd3 using voilá-GAL4 in combination with Tubulin-GAL80ts and R57C10-GAL80 (EEC>) (N=4-6). c, Confocal imaging of upd3-GAL4 driven UAS-GFP expression in EECs, co-stained with Prospero (Pros) as an EEC marker (Scale bar, 20 µm). d, Twenty-four-hour sleep profiles in animals with EEC-specific upd2 and upd3 knockdown (N=29-34). e, Total daytime sleep and f, nighttime sleep durations in animals with EEC-specific upd2 and upd3 knockdown flies (N=29-34). g, Motion bout length for control and EEC-specific upd2 and upd3 knockdown flies (N=23-27). h, Motion bout activity across experimental groups (N=23-30). i, Twenty-four-hour sleep profiles for global upd2/3 mutants (N=19-32). j, Daytime and k, nighttime sleep durations in global upd2/3 mutants (N=19-32). l, Sleep profiles following EEC-specific CRISPR-mediated upd2 and upd3 knockout (N=24-31). m, Daytime and n, night sleep durations in animals with EEC-specific CRISPR-mediated upd2 and upd3 activation (N=25-31). Statistical analyses performed using Mann-Whitney tests for panels a and b; Ordinary one-way ANOVA with Dunnett’s multiple comparisons test for panels e, g, j, and k; Kruskal-Wallis One-Way ANOVA with Dunn’s multiple comparison test for panels f, h, m, and n. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

We then explored whether these EEC-derived cytokines govern sleep under such normal conditions, given the known role of cytokines in both healthy and inflammatory states. We found that the knockdown of the two main immune-related cytokines, upd2 or upd3 using voilà> increased the amount of time animals spent asleep (defined as a period of inactivity lasting at least five minutes), especially during the day (Fig. S1a-c). This phenotype was reproducible using the more restricted EEC> driver that includes R57C10-GAL80, suggesting that the EEC-specific loss of these cytokines promotes sleep with pronounced effects during daytimes when females are typically active (Fig. 1d-f). This outcome was not attributable to off-target effects, as two independent RNAi lines targeting the main IL-6-like cytokine, upd3, produced a similar phenotype or effects of the RNAi transgenes themselves (Fig. 1d-f and S1d-f). Animals lacking EEC-derived Unpaired signaling also exhibited shorter motion bouts (periods of activity; Fig. 1g and S1g). However, the EEC-specific knockdown of upd2 or upd3 did not reduce motion-bout activity (the intensity of activity during wake periods), implying that the lack of gut cytokine signaling did not reduce general activity when the animals were awake (Fig. 1h and S1h), and these effects do not arise from RNAi transgene insertions themselves (Fig. S1i,j). These observations suggest a direct influence of gut-derived Upd2 and Upd3 on sleep rather than a broader impact on general activity levels. Given that sleep and feeding are mutually exclusive behaviors, we measured feeding. We did not detect significant alterations in feeding behavior as a consequence of upd2 or upd3 knockdown in EECs over a 24-hour period using the automated FLIC system (Fig. S1k), nor did we observe an effect of EEC-specific upd3 overexpression on food consumption via a dye assay (Fig. S1l). Furthermore, EEC-specific upd3 knockdown did not affect the animals’ metabolic state as reflected in their levels of stored triacylglyceride (TAG) (Fig. S1m). Therefore, the sleep phenotype exhibited by animals with EEC-specific upd2/3 knockdown is not associated with changes in metabolism, appetite or feeding behavior.

To further explore the role of Upd2 and Upd3 in sleep regulation, we made use of deletion mutations that disrupt either both upd2 and upd3 (upd2,311) or upd3 alone (upd311). Both mutants exhibited a pronounced increase in sleep (Fig. 1i-k), with strong effects on daytime sleep phenocopying the RNAi-mediated knockdown in the EECs. We additionally disrupted the upd2 or upd3 genes specifically in the EECs using somatic tissue-specific CRISPR-mediated deletion. EEC-specific CRISPR-mediated knockout of upd2 or upd3, induced by UAS-controlled gRNA pairs designed to excise portions of each gene’s coding sequence, led to significantly elevated sleep (Fig. 1l-n), further reinforcing these cytokines’ role in sleep modulation. Collectively, these findings demonstrate that Upd2 and Upd3 expressed by EECs are important modulators of diurnal sleep patterns in Drosophila under normal homeostatic conditions.

Glial cytokine JAK-STAT signaling regulates sleep

To identify the CNS targets of EEC-derived Upd2/3 cytokine signaling by which they regulate sleep, we examined the effects of targeted knockdown of the Upd2/3 JAK-STAT-linked receptor dome in neurons or glia, the two main cell types in the CNS. Pan-neuronal dome knockdown using the driver R57C10-GAL425 did not significantly alter daytime sleep in adult females (Fig. 2a), contrary to the sleep increase observed upon loss of upd2 or upd3 in EECs (Fig. 1). This suggests that neurons are not the targets by which gut Unpaired cytokine signaling regulates sleep. In contrast, knockdown of dome in glial cells using the pan-glial driver repo-GAL4 (repo>) resulted in a pronounced increase in daytime sleep (Fig. 2a), similar to the phenotype observed upon EEC-specific loss of Unpaired cytokines. To substantiate this observation further, we silenced dome expression in glia using three independent RNAi lines, which all increased sleep with strong induction of sleep during the day, effectively ruling out any off-target or transgene-background effects (Fig. 2b-d and S2a-e). Additionally, animals with glial dome knockdown displayed shorter daytime motion-bout lengths, suggesting reduced periods of wakefulness (Fig. 2e and S2f), without exhibiting decreased activity during these bouts (Fig. 2f and S2g), phenocopying the effects of upd2/3 knockdown in EECs (Fig. 1g,h). We also assessed whether glial-specific dome knockdown might affect feeding and energy storage, but we observed no reduction in food intake (Fig. S2h), and no changes in TAG levels were detected (Fig. S2i). These findings collectively argue that augmented sleep resulting from impaired JAK-STAT signaling in glia is not due to a general decline in activity but rather represents a specific regulation of sleep itself.

EEC-derived Unpaired signaling regulates glial JAK-STAT activity that modulates sleep.

a, Day sleep measurements for flies with knockdown of IL-6 related Unpaired cytokine receptor domeless (dome), which activates JAK-STAT, in neurons (R57C10-GAL4, R57C10>) and glial cells (repo-GAL4, repo>) (N=25-39). b, Twenty-four-hour sleep profiles for controls and animals with glial-specific dome knockdown (N=25-32). c, Total day sleep duration and d, total night sleep duration for animals with glial-specific dome knockdown and control flies (N=25-32). e, Motion bout length and f, motion bout activity in animals with glial-specific dome knockdown and controls (N=25-32). g, Images of brain from controls and animals with EEC knockdown upd2 or upd3 using voilá>10xSTAT-GFP. STAT-GFP expression (green) with Repo labeling (red) indicates glial cells (Scale bar, 50 µm). Insets show zoomed views of STAT-GFP+ and Repo+ glial cells (Scale bar, 15 µm). h, Quantitative analysis of GFP intensity in the layer of glial cells located at the surface of the brain in animals with EEC knockdown of upd2 or upd3 and controls (N=8). Statistical analyses were conducted using parametric t-tests for panel a; Kruskal-Wallis One-Way ANOVA with Dunn’s multiple comparison test for panels c, d, e and f; and Ordinary one-way ANOVA with Dunnett’s multiple comparisons test for panel h. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

To directly assess the functional Unpaired-mediated communication between the gut and glial cells, we manipulated upd2 and upd3 in EECs of animals carrying a ubiquitous transgenic JAK-STAT reporter (10xSTAT-GFP)40. Knockdown of either upd2 or upd3 in the EECs led to a marked decrease in JAK-STAT reporter activity within Repo-positive glial cells under normal conditions (Fig. 2g,h), suggesting that Upd2/3-mediated signaling from the EECs to the brain’s glial cells activates JAK-STAT signaling. Taken together this suggests that gut-to-glia communication via Upd2 and Upd3 modulates diurnal sleep patterns through glial JAK-STAT activation and that these cytokines are required for the maintenance of wakefulness during the day under healthy conditions.

Oxidative stress modulates sleep through gut-derived cytokine signaling

Having established the significance of gut-derived Unpaired cytokines in maintaining wakefulness under normal conditions, we next explored their role in sleep-regulation during gut disturbances that trigger immune and inflammatory responses. Infection or inflammation in the gut leads to increased levels of ROS, causing local intestinal oxidative stress26,27. To probe the effects of enteric ROS on sleep and cytokine signaling, we induced oxidative stress in the gut by feeding animals food containing 1% hydrogen peroxide (H2O2) for 20 hours. We asked whether intestinal oxidative stress might elevate the levels of Upd2 and Upd3 in the gut and found substantial upregulation of upd3 expression in ROS-challenged dissected midguts of adult females (Fig. 3a). This effect mirrored the upregulation observed with EEC-specific overexpression of upd3, indicating that it reflects physiologically relevant production of Upd3 by the gut in response to oxidative stress. Although oxidative stress also promoted upd2 expression, albeit to a lesser extent, this effect was not modulated by simultaneous EEC-specific upd3 overexpression.

EEC-derived Unpaired signaling modulates sleep in response to intestinal oxidative stress.

a, Measurement of upd2 and upd3 expression in the midguts upon 20-hour treatment with 1% H2O2-laced food or overexpression of upd3 (upd3-OE) using EEC> (N=4). b and c Assessment of sleep duration over consecutive days during daytime (ZT0-ZT12) and nighttime (ZT12-ZT24) in animals exposed to food containing (b) 0.1% H2O2 (N=26-31) or (c) 1% H2O2 (N=23-29). d, Daily sleep and e, nightly sleep amounts measured over one day under standard food conditions followed by two consecutive days on 1% H2O2-containing food in animals with EEC-specific upd2 or upd3 knockdown and controls (N=23-30). Experiments measuring sleep levels in controls and animals lacking EEC-derived upd2 or upd3 were performed concurrently and share the control but are presented in separate figures (b, c, d, and e) for clarity. f, Survival rates under a 1% H2O2-induced oxidative stress diet in controls and animals with EEC-specific upd2 or upd3 knockdown (N=23-30). g, A 48-hour sleep profile comparison between global upd2/3 mutants and w1118 controls under one day of standard food conditions followed by one day of 1% H2O2-induced stress (N=18-63). h and i, Quantification of daytime and nighttime sleep durations in upd2/3 mutants versus w1118 controls under normal food conditions and the following day exposed to food containing 1% H2O2 (N=18-63). j, Observation of sleep patterns and k, measurements of daytime sleep across a three-day period, encompassing a day on standard diet, subsequent day on 1% H2O2-laced food to induce oxidative stress, and a final day back on standard diet to monitor recovery, in flies with EEC-specific overexpression of upd3 (upd3-OE) compared to control (N=29-32). Statistical analyses were performed using Kruskal-Wallis One-Way ANOVA with Dunn’s multiple comparison test for panels a, b, c, d, e, f, and k; Mann-Whitney test for panels h and i. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

We next investigated whether sleep is modulated by intestinal oxidative stress and if Unpaired signaling from EEC is required for this response. We induced intestinal oxidative stress by exposing animals to a diet supplemented with H2O2 at Zeitgeber Time 0 (ZT0), the onset of the light phase in a 12-hour light/dark cycle. Exposure to a lower H2O2 concentration (0.1%) incrementally increased daytime sleep amount over successive days (Fig. 3b). In contrast, a higher H2O2 concentration (1%) triggered an immediate augmentation of daytime sleep (Fig. 3c). These observations suggest dose-dependent modulation of sleep by intestinal oxidative stress. Since 1% H2O2 feeding induced robust responses both in upd3 expression and in sleep behavior, we asked whether gut-derived Unpaired signaling might be essential for the observed ROS-induced sleep modulation. Indeed, EEC-specific RNAi targeting upd2 or upd3 abolished the sleep response to 1% H2O2 feeding. These animals with EEC-specific knockdown of upd2 or upd3 did not exhibit increased daytime sleep in response to the induction of oxidative stress in the intestine, even over two consecutive days exposed to 1% H2O2-containing diet (Fig. 3d,e). The specificity of this response was corroborated by three independent RNAi lines targeting upd3, negating the possibility of RNAi off-target effects (Fig. 3d,e), and the loss of response to ROS was also not attributable to the transgenes themselves (Fig. S3a-d). Intriguingly, animals lacking upd3 in the EECs not only did not increase their sleep under oxidative stress but indeed appear to lose nighttime sleep in response to enteric stress.

We next tested whether this sleep phenotype might be associated with general physiological processes rendering animals lacking EEC Unpaired signaling more susceptible to ROS-induced damage. However, when we assessed survival on 1% H2O2-containing food, animals with upd2 or upd3 knockdown in EECs displayed no additional sensitivity to oxidative stress, compared to controls (Fig. 3f). This lack of effect indicates that the loss of EEC-derived Unpaired signaling specifically leads to an impaired behavioral sleep response to intestinal oxidative stress, rather than to compromised physiological processes that would make the animals more vulnerable to oxidative stress insults. We also examined whether animals lacking EEC-derived Unpaired signaling exhibit normal behavioral responses to other conditions that modulate sleep, which would suggest a specific requirement for this signaling in responding to intestinal oxidative stress. Animals typically suppress their sleep in response to nutritional deprivation, a behavior conserved across species that is believed to facilitate food-seeking activities and that is also influenced by EEC-mediated hormone signaling25,41. Animals lacking EEC-derived upd3 suppressed their sleep similar to controls in response to starvation, indicating a normal sleep response to nutritional stress (Fig. S3e-g).

Although we observed behavioral phenotypes with manipulations of either upd2 or upd3, suggesting that both are required for normal function, Upd2 and Upd3 likely function at least partially redundantly or additively in their regulation of sleep, as is the case for other processes42. Moreover, RNAi effects do not result in a complete loss of function. Therefore, we speculated that a stronger disruption and combined knockout of both upd2 and upd3 might lead to even more pronounced phenotypes. We therefore tested the upd3Δ deletions and the double upd2,3Δ double deletion mutants. Whereas upd3Δ and upd2,3Δ mutants exhibited increased baseline sleep under homeostatic conditions, these animals not only failed to increase their sleep in response to oxidative stress but indeed showed a strong reduction in daytime and nighttime sleep under oxidative-stress conditions (Fig. 3g-i). These results suggest that contrary to its role in promoting wakefulness during normal homeostatic conditions, the enhanced ROS-induced Unpaired signaling from EECs helps sustain a higher sleep level during periods of oxidative stress. This indicates a dual functionality of Unpaired cytokine signaling, in which low Unpaired signaling promotes wakefulness under normal conditions, whereas higher ROS-induced Unpaired signaling facilitates a shift to restorative sleep during intestinal stress.

We therefore investigated whether higher levels of Unpaired signaling from the gut, comparable to the level produced during oxidative stress, could enhance sleep in the absence of exogenous stressors. We analyzed the effect of upd3 overexpression in EECs, which drives expression of midgut upd3 to levels similar to those induced by 1% H2O2 feeding (Fig. 3a). Consistent with a model in which high levels of Upd3, which would occur during high levels of intestinal oxidative stress, promotes daytime sleep, animals overexpressing upd3 in the EECs exhibited increased sleep during the day, even in the absence of H2O2-induced oxidative stress (Fig. 3j,k). These animals further increased their sleep in response to H2O2-induced enteric oxidative stress, unlike those lacking gut-derived upd3 (Fig. 3d,e), suggesting they are able to mount an additional ROS-induced Unpaired signaling response on top of the overexpression-induced levels. After animals were switched back to normal food after one day of oxidative stress, both control animals and those with EEC-specific upd3 overexpression exhibited even more sleep than during the previous day under oxidative-stress conditions (Fig. 3j,k). This suggests a robust recovery sleep response following the insult, likely mediated by Unpaired signaling, since the effect is more pronounced with upd3 overexpression. Taken together our results show that control animals increase their sleep during oxidative stress, likely as an adaptive recovery response. In contrast, animals with EEC-specific knockdown of unpaired cytokines do not exhibit this ROS-induced sleep response; instead, they experience sleep loss during such stress conditions. This suggests that while keeping the animal awake during normal healthy conditions, ROS-induced gut Unpaired signaling temporarily suppress these wake-promoting effects, leading to more sleep.

EEC-derived Unpaired cytokine signaling activates glial JAK-STAT under oxidative stress

To investigate whether oxidative stress enhances glial JAK-STAT signaling and if this enhancement is mediated by gut-derived Upd2 and Upd3, we assessed glial JAK-STAT reporter activity using the dual-color TransTimer system, which provides temporal information about JAK-STAT signaling43. In this system, active STAT promotes the expression of a construct encoding a short-lived destabilized GFP (dGFP, half-life ∼2 hours) and a long-lived RFP (half-life ∼20 hours) separated by a 2A peptide (6xSTAT-dGFP:2A::RFP); a higher ratio of GFP to RFP in a given cell reflects more recent JAK-STAT signaling. We explored whether JAK-STAT signaling responds dynamically to intestinal oxidative stress and assessed two daily time points. We observed no circadian changes between ZT0 (lights on) and ZT12 (lights off) in the superficial layer of cells surrounding the brain (Fig. 4a,b), which is composed of glia44. However, we observed a significant increase in GFP signal at ZT0 in animals fed for 20 hours with 1% H2O2-containing food, indicating recent JAK-STAT activity in the surface glia. Next, we investigated whether gut-derived Unpaired signaling is responsible for this upregulation by combining the 10xSTAT-GFP reporter with knockdown of upd2 or upd3 in the EECs. Whereas glial JAK-STAT reporter activity was upregulated by oxidative stress (20 hours 1%-H2O2 feeding) in control animals, this response was abolished in animals with EEC-specific knockdown of upd2 or upd3, indicating that this response is dependent on these EEC-derived cytokines (Fig. 4c,d). Since in this case we used voilá> without pan-neuronal R57C10-GAL80 to drive knockdown in EECs, we measured the expression of upd2 and upd3 in heads to check for any unintended neuronal effects that might contribute to the observed effect on glial JAK-STAT activity. We detected no changes in the expression of these genes in the heads, confirming that the observed JAK-STAT activation in glial cells is attributable to cytokines derived from EECs (Fig. S4a,b). To test the ability of gut-derived Upd3 to drive events in the brain in another way, we made use of cells’ homeostatic response to changes in signaling input. Receptor expression is often upregulated in response to low levels of a ligand as a compensatory mechanism to enhance cellular sensitivity45. We observed an upregulation of dome transcript levels in the heads of animals with EEC-specific knockdown of upd3, exposed to oxidative stress induced by 15 hours of feeding with food laced with 1% H2O2 (Fig. 4e). Increased dome expression suggests reduced Unpaired ligand availability as a result of the loss of EEC-derived Upd3. Together, our results demonstrate that EEC-derived Unpaired cytokine signaling is required for activating glial JAK-STAT under oxidative stress.

Activation of glial JAK-STAT by EEC-derived Unpaired cytokines in response to oxidative stress.

a, Representative images of brains from flies expressing STAT::dGFP:2A::RFP reporter, with green representing destabilized GFP (dGFP) and purple indicating stable RFP. Left panels show brains under normal light conditions (ZT0), middle panels show brains with lights off (ZT12), and right panels depict brains after 20 hours of oxidative stress induced by 1% H2O2-containing food, imaged at light-on time ZT0. White dotted lines outline the brain perimeter. Scale bar, 50 µm. b, Ratio of dGFP to RFP fluorescence intensity under normal light conditions at ZT0, at ZT12, and after oxidative stress (ZT0), as depicted in panel a, to show dynamic changes in JAK-STAT activity (N=108-461, indicating the number of cells counted). c, Representative images of brains displaying 10xSTAT::GFP expression under homeostatic conditions and after oxidative stress in control flies (Ctrl) and flies with EEC-specific upd2 or upd3 knockdown. Scale bar, 50 µm. Inset ’Zoom’ panels provide magnified views of glial cells labeled by Repo. Scale bar, 15 µm. d, Quantification of 10xSTAT::GFP intensity in glial cells under homeostatic and oxidative stress conditions, demonstrating the impact of EEC-specific cytokine knockdown (N=8, indicating the number of brains). e, qPCR analysis of dome expression in the brains of flies with EEC-specific upd3 knockdown in comparison to voilà> controls (N=5). Statistical analyses were conducted using Kruskal-Wallis One-Way ANOVA with Dunn’s multiple comparison test for panel b; ordinary one-way ANOVA with Tukey’s multiple comparison test for panel d; and t-tests for panel e. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

Glia JAK-STAT modulates sleep in response to oxidative stress

Since EEC-derived Unpaired signaling promotes oxidative stress-induced sleep and glial JAK/STAT activity, we investigated whether the observed glial JAK-STAT signaling is involved in the modulation of sleep in response to intestinal oxidative stress. Knockdown of dome in all glial cells using the repo> driver completely abolished the ROS-induced daytime sleep response when animals were fed 1% H2O2 supplemented diet. The specificity of this effect was confirmed using three distinct RNAi lines, and with transgenic RNAi controls (Fig. 5a and S5a-c). Similar to the effects observed with upd2,3 mutants (Fig. 3g-i), glia-specific dome knockdown (p < 0.00001 for dome-iKK and p = 0.0556 for dome-iTRiP) resulted in progressive and substantial sleep loss over two consecutive days on 1% H2O2-containing food. To rule out developmental effects, we restricted knockdown of dome in glial cells to the adult stage using the repo> driver in combination with Tub-GAL80ts (repoTS>) and observed similar effects (Fig. 5b).

EEC-derived Unpaired and glial Domeless signaling modulate sleep during intestinal oxidative stress.

a, Daytime sleep duration in flies with glial-specific dome knockdown under standard and oxidative stress conditions induced by 1% H2O2 in food (N=25-32). b, Daytime sleep duration in flies with repo-driven dome knockdown restricted to the adult stage using Tub-GAL80ts(repoTS>) under normal conditions and during exposure to 1% H2O2-containing food (N=29-32). c, Daytime sleep during a three-day period, encompassing a day on standard diet, subsequent day on 1% H2O2-laced food to induce oxidative stress, and a final day back on standard diet to monitor recovery, in controls and animals with glial-specific dome knockdown (N=24-32). d, Survival rates of controls and flies with glial-specific dome knockdown after exposure to oxidative stress by 1% H2O2-laced food (N=31). e, Sleep profiles and f, daytime sleep duration for animals with EEC-specific upd3 knockdown compared to control flies across a 36-hour period encompassing 24 hours on standard diet followed by 12 hours on oxidative stress conditions, induced by 4% H2O2-containing food (N=15-30). g, Sleep profiles and h, daytime sleep duration for animals with glial-specific dome knockdown compared to control flies across a 36-hour period encompassing 24 hours on standard diet followed by 12 hours on oxidative stress conditions, induced by 4% H2O2-containing food (N=20-32). i, Nighttime sleep durations for animals under 4% H2O2 oxidative stress with adult restricted knockdown of dome in glial cells and control (N=31-32). Statistical tests: Kruskal-Wallis One-Way ANOVA with Dunn’s multiple comparison test for panels a, c and d; Unpaired t-tests for panels b, f, and h; and Mann-Whitney test for panel i. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

We then examined the dynamics of glial sleep-regulatory JAK-STAT signaling by inducing oxidative stress for one day and then transferring the animals to normal food to observe the recovery response. The results showed that, in response to 1% H2O2-containing food, animals lacking glial dome expression displayed a sleep response opposite from that of controls, with a reduction in sleep duration rather than an increase, confirmed using independent RNAi lines (Fig. 5c). This phenotype is similar to that seen in the upd2,3Δ double mutants (Fig. 3h-j). During the recovery phase, after the animals had been switched back to normal food, the sleep level of controls increased even further, but the sleep duration exhibited by animals with glia-specific dome knockdown immediately reverted to pre-stress levels. This pattern indicates that animals with inhibited glial JAK-STAT signaling display an aberrant dynamic sleep response to oxidative stress that is not a consequence of a physiological breakdown but rather arises from altered inhibitory sleep-regulating mechanisms. In line with this and similarly to the loss of upd2 or upd3 in the EECs, dome knockdown in glial cells did not decrease survival on H2O2-containing food (Fig. 5d). This supports the notion that physiological resistance to oxidative stress remains unaltered by gut-glial Unpaired signaling, which in turn indicates that the signaling modulation leads to a specific sleep phenotype. We further assessed whether loss of glial dome affected sleep responses induced by sleep deprivation by evaluating the animals’ ability to recover sleep after deprivation occurring during the second half of the night (ZT18-ZT24). Like controls, animals with glial-specific dome knockdown exhibited increased sleep (rebound sleep) in the morning hours (ZT0-ZT2) following sleep deprivation (Fig. S5d,e). This indicates that they exhibit normal rebound sleep responses to deprivation and retain the capability to further increase their sleep. Collectively, these data suggest that Dome-mediated JAK-STAT signaling in the glial cells specifically regulates ROS-induced sleep responses.

Furthermore, we investigated whether increased intestinal oxidative stress would exacerbate the phenotypes associated with the loss of upd3 in EECs or dome in glial cells by exposing the animals to food containing 4% H2O2 and observing sleep-architecture changes. Oxidative stress resulted in increased sleep in control animals as anticipated (Fig. 5e-h). However, in animals with EEC-specific upd3 knockdown, or glia-specific dome RNAi, exposure to 4% H2O2-containing food led to a pronounced loss of sleep during the daytime. For the EEC-specific upd3 knockdown, the RNAi effect was induced at the adult stage (Fig. 5e,f). We therefore also confirmed that adult-restricted knockdown of dome in glial cells resulted in similar phenotypes (Fig. 5i). Thus, under conditions of intensified intestinal stress induced by 4% H2O2 in the food, the loss of upd3 in EECs phenocopies the glial knockdown of dome, leading to reduced sleep and increased wakefulness.

BBB glial JAK-STAT pathway activation drives sleep in response to intestinal oxidative stress

To determine the subset of glial cells responsible for mediating Unpaired-driven sleep regulation, we focused on the perineurial and subperineurial glial cells that form the BBB. These BBB glial cells serve as the interface between the CNS and the periphery, including its organs44, and are ideally situated to receive circulating signals from the intestine. Using the 10xSTAT-GFP reporter, we assessed whether Upd3 from the gut activates JAK-STAT signaling within BBB glial cells. Knockdown of upd3 in the EECs using voilá> (without R57C10-GAL80) which drives specific knockdown in the gut without any neuronal effects (Fig. S4a,b), resulted in decreased GFP intensity in the outermost glial cell layer of the central brain facing the periphery after 15 hours of exposure to 1% H2O2-containing food, indicating reduced JAK-STAT activity in these cells (Fig. 6a,b). Interestingly, this knockdown did not affect JAK-STAT activity in the outer glial layer of the ventral nerve cord (VNC), suggesting that Upd3 acts specifically on brain BBB glia in response to intestinal oxidative stress.

BBB glia drives Domeless-mediated sleep responses to intestinal oxidative stress.

a, Representative images showing GFP expression driven by 10xSTAT::GFP in controls and animals with voilà-GAL4 (voilà>) driven upd3 knockdown in EECs under intestinal ROS induced by 20 hours exposure to 1% H2O2-containing food. The top panels depict overall brain and ventral nerve cord (VNC) structure with views of surface or deeper layers; the bottom panels provide zoomed-in views, highlighting the BBB glia at the interface between the brain and external environment. Dotted lines indicate brain and VNC perimeters. Scale bars, 50 µm (top) and 15 µm (bottom). b, Quantification of GFP intensity in the brain and VNC in BBB glial controls and animals with EEC knockdown of upd3 under ROS stress, induced by exposure to 1% H2O2-laced food (N=7). c, Daytime and d, nighttime sleep durations in flies with BBB glial-specific knockdown of dome or overexpression of hopTum under normal conditions, during exposure to 1% H2O2-containing food, and subsequent recovery on normal diet (N=23-32). e, Daytime and f, nighttime sleep durations in flies with BBB glial-specific knockdown of dome or overexpression of hopTum under normal conditions, during exposure to 4% H2O2-containing food, and subsequent recovery on normal diet (N=23-32). Statistical tests used: Unpaired t-tests for panel b; Kruskal-Wallis One-Way ANOVA with Dunn’s multiple comparison test for panels c, d and f; Ordinary one-way ANOVA with Dunnett’s multiple comparisons test for panel e. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

We further examined BBB-specific JAK-STAT signaling effects on sleep by driving expression in subperineurial glial cells – those that form the permeability barrier – with moody-GAL4 (moody>). As observed with other manipulations, knockdown of dome in these BBB glial cells led to increased sleep during normal homeostatic conditions (Fig. 6e,f). Whereas controls exhibited the expected sleep increase both during 1%-H2O2 exposure and during the recovery period, loss of dome in the subperineurial BBB glia blocked these effects. When oxidative stress levels were elevated further using 4% H2O2-containing food, dome-RNAi in BBB glial cells led to sleep loss in response to oxidative stress, with sleep levels rebounding to pre-stress levels on the subsequent recovery day when animals were returned to a normal diet (Fig. 6e,f). These results suggest that disrupting dome specifically in BBB glial cells recapitulates the phenotypes observed with pan-glial knockdown or EEC-specific upd3 knockdown, indicating that Dome-mediated JAK-STAT activation in subperineurial BBB cells is required for maintaining an increased sleep state during intestinal oxidative stress.

To assess the sufficiency of subperineurial JAK-STAT signaling in inducing sleep, we activated the pathway in these cells by expressing a hyperactivated variant of the Drosophila JAK ortholog hopscotch (hopTum). Activation of JAK in BBB glia increased sleep under normal conditions without oxidative stress, consistent with a sleep-promoting effect of high JAK-STAT signaling (Fig. 6c-f). Moreover, animals with overactivate JAK signaling in BBB glia exhibited further enhanced sleep both during oxidative stress and in the subsequent recovery phase, as opposed to those with dome knockdown. This suggests that the combined activation of JAK-STAT induced by intestinal ROS and expression of hopTum leads to additive increases in sleep. Collectively, our data indicate that JAK-STAT signaling specifically in the subperineurial glial of the BBB links sleep responses to intestinal oxidative stress.

AstA signaling promotes wakefulness and mediates ROS-induced sleep regulation in BBB glia

Our results indicate that the effect of gut-to-glial Unpaired cytokine signaling is both dose-dependent and context-dependent. During intestinal oxidative stress, ROS-induced EEC Unpaired signaling leads to high JAK-STAT activity in subperineurial glial cells. Given that animals lacking this gut cytokine-to-glial signaling fail to maintain a high sleep state during oxidative stress, instead exhibiting increased wakefulness, this pathway appears to suppress wake-promoting signals under such conditions. The role of such wake-suppressive effects is likely to enhance sleep, aiding the process of recovery from intestinal damage. To identify potential wake-promoting signals that might be gated by JAK-STAT signaling, we examined a published dataset of genes whose expression in glia is positively or negatively correlated with these cells’ JAK-STAT activity following enteric infection33. The receptors for Allatostatin A (AstA), AstA-R1 and AstA-R2, both ranked among the top 4% of genes most strongly downregulated by JAK/STAT signaling (with AstA-R1 expression reduced by ∼80% and AstA-R2 by ∼90%). Notably, these were the only peptide-hormone G-protein-coupled receptors (GPCRs) downregulated in the JAK-STAT-activate glial cells. This suggests that upon intestinal infection, the JAK-STAT pathway is activated in glial cells, which suppresses AstA signaling by reducing the expression of the AstA receptors. Considering the central role of neuronal AstA in sleep-regulatory circuits46,47, we investigated whether AstA might constitute a wake-promoting signal that is inhibited in glial cells by gut-derived Unpaired signaling. To evaluate the expression pattern of AstA-R1 and AstA-R2 within glial populations, we employed AstA-R1-GAL4 and AstA-R2-GAL4 knock-in constructs to drive the expression of nuclear-localized RFP. We co-stained the brain with antibodies against the glial transcription factor Repo, which marks the nuclei of glial cells. We observed that the outer layer of glial cells at the barrier between the brain and the periphery – constituting the BBB – express both AstA-R1 and AstA-R2 (Fig. 7a). These findings are in line with previously reported data showing expression of AstA-R1 and AstA-R2 in glial cells referred to above33. To functionally characterize the role of AstA signaling in these cells, we knocked down AstA-R2 in BBB glial using moody>. This led to a significant reduction in AstA-R2 transcript levels in dissected brains, indicating that BBB glia is a significant source of AstA-R2 expression (Fig. S6a). In support of an inhibitory role of Unpaired signaling, we observed that AstA-R1 and AstA-R2 expression was upregulated in the heads of animals with EEC-specific upd3 knockdown, following gut-oxidative stress induced by feeding with 1% H2O2-laced food for 20 hours (Fig. 7b). To demonstrate that this is caused by a failure to suppress AstA receptors in glial cells, we examined AstA receptor expression in brains following glia-specific dome knockdown in animals fed 1% H2O2-containing food for 20 hours. Indeed, glial-specific dome knockdown led to strong upregulation of both AstA-R1 and AstA-R2, indicating that ROS-induced glia-mediated Unpaired signaling is inhibiting AstA receptor expression (Fig. 7c).

Gut Unpaired cytokine signaling inhibits wake-promoting AstA signaling.

a, Images of brains from animals with AstA-R1-GAL4 and AstA-R2-GAL4 driving nuclear dsRed (nRFP, magenta) and co-stained with anti-Repo antibodies (green) show glial cells. The yellow dashed line indicates the interface between the brain and the external space, with the area below housing the (BBB glial cells (Scale bar, 10 µm). The yellow demarcation accentuates the separation between the cerebral interior and the external milieu, identifying the location of BBB glial cells underneath this partition (Scale bar is 10 µm). b, Relative expression of AstA-R1 and AstA-R2 in heads from animals with upd3 knockdown in EECs driven by voilà-GAL4 (voilà>) compared to the control group after 20 hours on 1% H2O2-laced food to induce oxidative stress (N=5-6). c, Relative expression of AstA-R1 and AstA-R2 in brains from animals with dome knockdown in glial cells driven by repo-GAL4 (repo>) compared to the control group after 20 hours on 1% H2O2-laced food to induce oxidative stress (N=6).d Sleep patterns and e, daytime sleep across a three-day period, encompassing a day on a standard diet, a subsequent day on 1% H2O2-laced food to induce oxidative stress, and a final day back on the standard diet to monitor recovery, in flies with BBB glial-specific knockdown of AstA-R2 compared to control (N=28-32). f, AstA transcript levels in brains and midguts from controls and animals with knockdown of AstA in AstA+ EECs using AstA-GAL4 (AstA>) in combination with R57C10-GAL80 to suppress neuronal GAL4 activity, referred to as AstAGUT> (N=5-6). g Sleep profiles and h, daytime sleep on standard food of animals with AstA knockdown in AstA+ EECs using AstAGUT> and controls (N=30-32). i Sleep profiles, j daytime sleep on standard food of controls and animals with TrpA1-mediated activation of AstA+ EECs, and animals with TrpA1-mediated activation of AstA+ EECs with simultaneous knockdown of AstA (N=30-32). k, Diagram illustrating the role of enteroendocrine cells (EECs) in regulating wakefulness and sleep through Unpaired cytokine signaling under homeostatic and disease conditions. Left: Under homeostatic conditions, EECs release baseline levels of Unpaired, which interacts with the blood-brain barrier (BBB) to maintain normal JAK-STAT signaling, promoting healthy wakefulness. Right: In response to disease, reactive oxygen species (ROS) increase in EECs, leading to elevated release of Unpaired. This surge in Unpaired upregulates JAK-STAT signaling in BBB glia, which inhibits wake-promoting AstA signaling by suppressing AstA receptors, thus resulting in increased sleep, a state termed ’sickness-sleep,’ to promote recovery. The diagrams depict the gut lining with EECs highlighted, the interface with the BBB, and the resulting systemic effects on the organism’s sleep-wake states. EC; enterocyte. Statistical tests used: Unpaired t-tests for panel b, c and f; Ordinary one-way ANOVA with Dunnett’s multiple comparisons test for panel e, h and j. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

Next, we investigated whether AstA receptors are involved in mediating the glial-regulated sleep response to intestinal oxidative stress. Like the Unpaired cytokines, AstA is released from EECs in response to oxidative stress48. It may therefore act as a context-dependent wake-promoting signal, that under certain conditions, is inhibited by Unpaired signaling in BBB glial to promote sleep. We thus hypothesized that during intestinal disturbances characterized by oxidative stress, gut-derived Unpaired signaling via JAK-STAT activation either sustains or consolidates sleep through a mechanism that involves the downregulation of wake-promoting AstA receptor signaling in BBB glial cells. In this model, EEC-derived Unpaired signaling normally suppresses AstA signaling in BBB glial cells under oxidative stress. Consequently, knocking down dome in these cells or unpaired in the EECs leads to a failure to downregulate AstA receptors, causing the animals to wake up under these conditions. Thus, inhibition of glial AstA receptors would impair the animals’ ability to respond to these wake-promoting signals altogether, leaving them unresponsive to intestinal ROS in terms of sleep. Consistent with this notion, we found that knocking down AstA-R1 or AstA-R2 in BBB glia attenuated the ROS-induced sleep response (Fig. S6b,c). Knockdown of AstA-R2 with second, independent RNAi line resulted in a more pronounced phenotype, with an almost completely blunted sleep response to intestinal ROS, showing no significant sleep increase during oxidative stress or the following day of recovery (Fig. 7d,e). AstA-R2 was also the most highly upregulated in response to loss of upd3 in the EEC or dome in glia (Fig. 7b,c), and it was the more strongly downregulated of the two AstA receptors in response to glial JAK-STAT activation, together suggesting that AstA-R2 is a primary receptor mediating these effects. Furthermore, knockdown of AstA receptors in BBB glia increased daytime sleep under normal homeostatic conditions, consistent with a wake-promoting role of AstA signaling in BBB glial (Fig. 7d,e and S6b,c).

AstA is produced by two cell types, neurons and the EECs in the gut46. Since BBB glial cells are well-positioned to receive hormonal signals from the periphery, they likely are regulated by gut-derived AstA. We thus examined whether gut-derived AstA acts as a wake-promoting signal, by conducting AstA knockdown in AstA-positive EECs using an AstA:2A::GAL4 knock-in in combination with R57C10-GAL80 to suppress GAL4 activity in the AstA-positive neuronal population. We confirmed that this driver efficiently reduces the expression of AstA in midguts without affecting neuronal AstA transcripts levels (Fig. 7f). Knockdown of AstA in AstA-positive EECs with either of two independent RNAi constructs led to increased sleep without any contribution of the transgenic insertion backgrounds (Fig. 7g,h and S6d,e), indicating that gut-derived AstA is indeed a wake-promoting factor. To assess whether EEC-derived AstA is sufficient to promote arousal, we employed the thermosensitive cation channel Transient Receptor Potential A1 (TrpA1)49 to induce hormonal release from AstA-positive EECs. Activation of these EECs suppressed sleep, an effect that was abolished by simultaneous AstA knockdown, supporting that the wake-promoting role of EEC-derived AstA (Fig. 7i,j). Taken together, our findings suggest that enteric oxidative stress induces the release of Unpaired cytokines, which activate the JAK-STAT pathway in subperineurial glia of the BBB. This activation leads to the glial downregulation of receptors for AstA, which is a wake-promoting factor also released by EECs. Gut-derived Upd signaling thereby gates the effect of AstA at the BBB and permits increased sleep during periods of intestinal stress. This process may aid in recovery and maintain overall organismal homeostasis.

Discussion

Intestinal inflammation and microbial imbalance are strongly associated with sleep disturbances and mental disorders such as anxiety and depression8,12–14. The influence of gut health on CNS-dependent behaviors is thought to be mediated by the gut-brain axis, comprised of diverse signals secreted by the gut that act on the brain to induce behavioral responses50. Whereas the regulation of feeding behavior by this axis has been extensively studied, leading to revolutionary approaches to medical weight loss and diabetes control, the role of gut-brain signaling in regulating sleep – a behavior affected by nearly all illnesses51,52– remains poorly defined. Sickness induces a state of sleepiness, which is believed to be a conserved adaptive response that promotes recovery by supporting energy conservation and efficient immune activity5,6. However, the exact mechanisms driving sickness-induced sleep remain largely elusive. We have demonstrated here that intestinal illness, through the release of interleukin 6-like Unpaired cytokines from endocrine cells of the Drosophila gut, regulates sleep via a glia-mediated pathway. This gut-to-glia communication promotes sleep during intestinal illness, presumably to facilitate the restorative sleep essential for both physical and mental health. Our findings provide mechanistic insight into how perturbations of gut health can influence sleep, potentially contributing to understanding the link between gastrointestinal disorders, sleep disturbances, and mental illnesses.

Cytokines, key mediators of immune and inflammatory responses, are thought to modulate sleep/wake cycles under disease conditions53. Interleukins and TNFα, cytokines induced during illness in mammals, have been suggested to promote sleep to aid recovery from disease. However, most of these effects have been attributed to the actions of cytokines produced within the CNS, leaving open the question of how diseases affecting other parts of the body can drive sleep responses. In Drosophila, sleep induced by immune responses is known to be influenced by the NFκB ortholog Relish in fat tissue54, and the neuronally expressed gene nemuri drives sleep and connects immune function with sleep regulation6. However, inter-organ signaling mechanisms by which intestinal disease or stress regulate sleep have not yet been described in either flies or mammals. Intestinal infections or inflammation lead to elevated levels of ROS in the gut, and our findings demonstrate that enteric oxidative stress in the gut triggers the production of Upd2 and Upd3 cytokines by hormone-secreting EECs. These gut-derived cytokines signal the state of the intestine to brain glial cells, including those of the BBB, and modulate sleep. This glia-mediated gut-to-brain signaling promotes wakefulness in healthy animals under normal conditions, while inducing sleep in response to oxidative stress in the intestine. This indicates a dual functionality, with low levels of gut Unpaired signaling promoting wakefulness, while higher disease-induced levels act oppositely to enhance sleep (Fig. 7k). A similar dose-dependent effect has previously been observed for interleukins in rats, in which injection of IL-1 into the CNS stimulates or inhibits sleep depending on the dose55.

Our results indicate that Unpaired signaling in subperineurial glial cells – those forming the BBB – activates the JAK-STAT pathway, and they suggest that this effect inhibits wake-promoting AstA signaling by downregulating AstA receptor expression. AstA and its receptors, which are orthologous with the mammalian Galanin signaling system, have been linked to the regulation of sleep, feeding, and metabolism46,56,57. Mammalian glia also express receptors for Galanin58, which also regulates sleep59,60, further underscoring a conserved role in sleep modulation across species. AstA-producing neurons induce sleep by releasing glutamate onto sleep-regulatory neuronal circuits, although recent findings also suggest a wake-promoting role for AstA signaling47. Irrespective of neuronal AstA, our experiments clearly show that AstA released from EECs of the gut acts as a wake-promoting signal and that activation of AstA receptor signaling in BBB glial cells induces wakefulness. This highlights the potential of peptide hormones to elicit different effects depending on their source tissue and thus their accessible target cells – whether they are produced by the gut outside the BBB or by the CNS inside the barrier. A similar phenomenon has been demonstrated for neuropeptide F39,61.

Drosophila exhibit conserved behaviors such as sleep, arousal/wakefulness, and anxiety-like responses4,62–67, and the EECs of the fly gut produce diverse hormones similar to those of mammals17–22, potentially influenced by diet, microbiota, and inflammatory responses. This makes Drosophila an excellent model for studying behaviors influenced by gut conditions through gut-brain signaling. Our findings suggest that the oxidative-stress level within gut tissues, which is modulated by intestinal bacteria and immune activity26,27, regulates sleep via EEC-derived Unpaired signaling, potentially explaining the observed links between gut microbiota and sleep disturbances in both flies and humans13,68. Furthermore, in mammals, conditions such as inflammatory bowel disease that are linked with oxidative stress69 are often associated with sleep and mental-health disturbances8,12,14. Our results imply that cytokines, including interleukin signaling from an inflamed or diseased gut, might be a mechanism by which intestinal illnesses affect sleep and mental health. Our findings raise the possibility that these cytokines may act on glial cells that integrate and relay these gut signals to brain sleep-regulatory circuits.

The neurons of the CNS are isolated from the circulatory system by the BBB70 that restricts the transmission of some hormonal and cytokine signals from the periphery to neurons within the brain. Our work suggests that the BBB receives AstA and Unpaired signaling from the periphery. Other reports indicate that Unpaired cytokines from tumors and from enterocytes also can activate JAK-STAT signaling in BBB glia cells in Drosophila33,71. One effect of glial JAK-STAT activity seems to be the alteration of BBB permeability71, raising the possibility that EEC-derived Unpaired signaling in BBB glia, directly or through AstA signaling, modulates sleep via regulation of BBB permeability, which has been linked to homeostatic sleep regulation36. Furthermore, the endocytotic activity of BBB glia, important for cellular transport and barrier function, has also been associated with sleep regulation37, and thus JAK-STAT-induced changes could regulate sleep through alterations in BBB trafficking. Another possibility is that JAK-STAT activity might regulate glial metabolic support for neuronal activity and in this way affect sleep patterns. In any case, our findings highlight the involvement of BBB glial cells in transmitting signals from the gut to the brain, adding another layer to our understanding of body-to-brain communication, which suggests that the BBB does more than protect the brain; it also responds to peripheral signals to modulate brain function, presenting an intriguing area for future research into gut-brain signaling.

Methods

Drosophila stocks and husbandry

Flies were cultured using a standard cornmeal-based formulation (82 g/L cornmeal, 60 g/L sucrose, 34 g/L yeast, 8 g/L agar, 4.8 mL/L propionic acid, and 1.6 g/L methyl-4-hydroxybenzoate) maintained at 25 °C with 60% relative humidity under a 12-hour light/dark cycle. Post-eclosion, flies were transitioned to an adult-specific, cornmeal-free diet (comprising 90 g/L sucrose, 80 g/L yeast, 10 g/L agar, 5 mL/L propionic acid, and 15 mL/L of a 10% methyl-4-hydroxybenzoate solution in ethanol)72 for 4-7 days prior to experiments. Adult mated females were used for all experiments. Flies were separated by sex one day prior to experimental procedures. Strains harboring the temperature-sensitive Tubulin-GAL80ts transgene were initially reared at 18 °C on cornmeal food and then switched to the adult diet for 3-4 days post-eclosion, still at 18 °C. Subsequently, they were incubated at 29 °C for five to seven days to activate RNAi expression in advance of the experiments. To ensure optimal conditions, the flies were provided with fresh food every three days. The following lines used in this study were sourced from the Bloomington Drosophila Stock Center (BDSC) at the University of Indiana: R57C10-GAL4 (#39171); UAS-upd3-RNAiTRiP (#32859); UAS-dome-RNAiTRiP (#53890); AstA::2A::GAL4 (#84593); AstA-R1::2A::GAL4 (#84709); AstA-R2::2A::GAL4 (#84594); UAS-AstA-RNAiTRiP (#25866); UAS-AstA-R2-RNAiTRiP (#67864); UAS-mCD8::GFP (#5137); Tub-GAL80ts (#7108); repo-GAL4 (#7415); moody-GAL4 (#90883); UAS-TrpA1 (#26263); and UAS-dsRed was extracted from (#8546). Additional fly lines were acquired from the Vienna Drosophila Resource Center (VDRC): control line w1118 (#60000, which is isogenic with the VDRC RNAi lines); UAS-upd2-RNAiSH (#330691); UAS-upd3-RNAiKK (#106869); UAS-upd3-RNAiGD (#27136); UAS-dome-RNAiKK (#106071); UAS-dome-RNAiGD (#36356); UAS-AstA-RNAiKK (#103215); UAS-AstA-R1-RNAiKK (#101395); UAS-AstA-R2-RNAiKK(#108648). The upd3Δ and upd2,3Δ deletion mutants were kindly provided by Bruno Lemaitre. UAS-upd3 and UAS-hopTum lines were gifts from David Bilder. The STAT-dGFP:2A::RFP line was generously supplied by Norbert Perrimon. The voilà-GAL4 strain was graciously provided by Alessandro Scopelliti. The R57C10-GAL80 transgene, situated on the X chromosome, was kindly donated by Ryusuke Niwa. The 10xSTAT::GFP line was a gift from Julien Colombani. To ensure uniformity in genetic background and to create control groups with an appropriate genetic background, all GAL4 and GAL80 lines used in this study were backcrossed to a w1118 line for multiple generations before being outcrossed with the genetic background specific to the RNAi, CRISPR, or overexpression lines to serve as controls in the experiments25.

Generation of tissue-specific CRISPR lines

To facilitate tissue-specific CRISPR-based disruption of the upd2 and upd3 loci, constructs were prepared containing two gRNA target sequences, flanked by efficiency-enhancing tRNA sequences. One construct was prepared for upd2, and two transgenes, targeting different genomic sites, were made for upd3. The upd2 construct was designed to delete the region encoding the secreted Upd2 protein. One upd3 construct should delete the initiator ATG codon, and the other – the one used in this work – deletes the second exon, which contains a significant portion of the coding sequence. Target-sequence cassettes were assembled by first cloning the tRNA insert from plasmid pCFD6 (Addgene #73915) between long oligos containing the gRNA target sequences using Q5 polymerase (New England Biolabs, #M0491S). The vector and the PCR products were then integrated using Gibson assembly (NEBuilder HiFi DNA Assembly Master Mix, New England Biolabs, #E2621S). Clones were sequenced to verify accuracy, and correct constructs were integrated into the fly genome at the attP2 site (chromosome 3L) by BestGene (Chino Hills, CA). The sequences used for cloning the upd2 and upd3 CRISPR constructs, with gRNA sequences indicated in bold, are shown in Supplementary Table 1.

Sleep, activity, and survival assays

The Drosophila Activity Monitoring System (TriKinetics, Waltham, MA) was employed to track sleep and activity patterns. Single flies aged six to eight days after eclosion were placed into glass tubes using light CO2 anesthesia. On one end, the tubes were sealed with a foam plug; on the other was placed a detachable 250-μL PCR tube containing 90 μL of feeding medium: either 5% sucrose in 1% agar/water, 5% sucrose mixed with various concentrations of H2O2 in 1% agar/water, or plain 1% agar/water for starvation conditions. All food media contained 0.5% propionic acid and 0.15% methyl-4-hydroxybenzoate to prevent microbial growth, with H2O2 being supplemented once the food had cooled to below 40 °C. Monitoring of the flies’ locomotor activity and sleep began at the beginning of the light cycle, after the animals had spent their first day in the tubes acclimating. Following an additional 24 hours on the standard 5%-sucrose diet, the PCR tubes were replaced with fresh ones containing H2O2 or starvation media at the lights-on transition when animals were awake, to avoid unnecessary disturbances to the animals. For recovery experiments, animals were switched back to a 5%-sucrose diet after 24 hours on H2O2-laced food. Periods of inactivity lasting 5 minutes or longer were recorded as “sleep.” In the sleep deprivation studies, the flies were placed in DAM monitors and subjected to mechanical stimulation, which was produced by attaching the monitors to a vortexer mounting plate (TriKinetics) and vibrating them for 2 seconds at the start of each minute throughout the 6-hour interval leading up to the lights-on time. Recovery sleep was assessed in flies that experienced a reduction of more than 60% in their typical sleep during the deprivation period, using their sleep patterns from the 24-hour period before the onset of sleep deprivation as a baseline. The occurrence of recovery sleep was specifically evaluated during the first two hours immediately following the sleep deprivation phase. For survival assays, flies were loaded into tubes filled with either plain 1% agar/water for starvation or 1% H2O2 in 1% agar/water to test oxidative stress resistance. The time of death was recorded upon the complete cessation of movement.

Feeding Assays

Short-term food consumption was quantified using a spectrophotometric dye-feeding assay73,74. All food intake experiments were conducted during the time of the normal morning meal, one hour after lights-on in a 12:12-hour dark/light cycle. Flies were transferred without anesthesia to food (90 g/L sucrose, 80 g/L yeast, 10 g/L agar, 5 mL/L propionic acid, and 15 mL/L of a 10% methyl-4-hydroxybenzoate solution in ethanol) supplemented with 0.5% erioglaucine dye (brilliant blue R, FD&C Blue No.1, Sigma-Aldrich, #861146) and allowed to feed for 1 hour. A control group of flies was provided with undyed food to establish the baseline absorbance levels of fly lysates. For each genotype, 1-2 flies per sample were homogenized in 100 μL phosphate buffer (pH 7.5) using a TissueLyser LT (Qiagen) with 5-mm stainless-steel beads. Homogenates were centrifuged at 16,000 g for 5 minutes, and 50 μL of the cleared supernatant was transferred to a 384-well plate. Absorbance was measured at 629 nm for erioglaucine using an Ensight multi-mode plate reader (PerkinElmer). Standard curves for dye were employed to correlate absorbance readings with the amounts of food consumed.

To assess feeding behavior, interactions with food were monitored over a 20-24 hour period using the Fly Liquid-Food Interaction Counter (FLIC) apparatus75. Drosophila Feeding Monitors (DFMs; Sable Systems) were placed in an incubator set to 25 °C (or 29 °C for strains carrying GAL80ts), maintaining 70% humidity under a 12:12-hour light/dark cycle. Each of the 12 DFM chambers was filled with a 10% sucrose solution, and individual flies were introduced in the afternoon following the morning meal. After several hours of acclimation, evening feeding activity was recorded. The following morning, at lights-on, the DFMs were refilled with fresh sugar solution, and data from the morning meal were collected. The feeding behavior was recorded using the manufacturer’s software and analyzed using R Studio with the provided package (https://github.com/PletcherLab/FLIC_R_Code).

Immunohistochemistry and confocal imaging

Adult midguts, brains, and VNCs were dissected in cold PBS and fixed for 1 hour at room temperature in 4% paraformaldehyde/PBS with gentle shaking. After a quick rinse with PBST (PBS with 0.1% Triton X-100, Merck #12298), the tissues were washed three times for 15 minutes each in PBST. The tissues were then blocked for 30 minutes at room temperature in PBST containing 5% normal goat serum (Sigma) and subsequently incubated overnight (or two days for CNS samples) at 4 °C with primary antibodies diluted in the blocking solution with gentle agitation. After removing the primary antibody solution, tissues were rinsed once and washed three times for 20 minutes each in PBST. Secondary antibodies diluted in PBST were applied, and tissues were incubated overnight at 4 °C, followed by three PBST washes and one PBS wash. The samples were then mounted on poly-L-lysine-coated slides (Sigma, #P8920) in Fluoroshield mounting medium with DAPI (Nordic Biosite, #GTX30920), and imaged on a Zeiss LSM-900 confocal microscope using a 20× air or 40× oil objective with Zen software. Image stitching was performed using the Stitching function of Zeiss Zen Blue 3.1, and analysis was conducted using the open-source FIJI/ImageJ software package76. All samples within each figure panel were dissected, stained, and imaged simultaneously using identical settings and reagents. To quantify STAT-dGFP:2A::RFP staining, dissected brains were fixed for 10 minutes in 4% paraformaldehyde at room temperature with agitation, rinsed once with PBST, mounted, and imaged immediately without antibody staining. For 10xSTAT::GFP quantification, brain and VNC samples were stained as described above, Z-stacks were projected in FIJI using the “sum” method, and Repo-positive cells at the surface were manually segmented to measure the raw integrated density with local background subtraction. For measuring 10xSTAT::GFP across the BBB, linear regions of interest were drawn through the glial layer perpendicular to the brain surface at the plane showing maximum brain size using FIJI’s line tool; GFP intensity was then quantified along these lines, and the peak was recorded for each transect. Antibodies used included 1:50 mouse anti-Repo (University of Iowa Developmental Studies Hybridoma Bank, #8D12), 1:20 mouse anti-Prospero (University of Iowa Developmental Studies Hybridoma Bank, #MR1A), 1:500 mouse anti-GFP (ThermoFisher, #A11120), 1:500 chicken anti-GFP (ThermoFisher, #A10262), 1:1000 rat anti-mCherry (used against Ds Red, ThermoFisher, #M11217), 1:500 Alexa Fluor 488-conjugated goat anti-mouse (ThermoFisher, #A11001), 1:500 Alexa Fluor 555-conjugated goat anti-mouse (ThermoFisher, #A21422), 1:500 Alexa Fluor 647-conjugated goat anti-mouse (ThermoFisher, #A-32728), 555-conjugated goat anti-rat (ThermoFisher, #A21434), and 488-conjugated goat anti-chicken (ThermoFisher, #A11039).

Triacylglyceride measurements

Triacylglyceride concentrations were determined following established methods72,77 using the Randox Triglycerides (GPO-PAP) method (Randox, #TR210). For each sample, flies were homogenized in 50 μl PBS per fly (between 2-4 flies per sample) containing 0.1% Tween-20 (Sigma #1379) using a TissueLyser LT (Qiagen) with 5-mm stainless-steel beads, 50 oscillations/second for 30 seconds. Homogenates were heated at 70 °C for 10 minutes to inactivate endogenous enzymes and centrifuged at 11,000 g for 1 min. Aliquots of cleared and vortexed supernatants (4 μl) were added to 36 μl of triglyceride reagent (Randox, #TR210) in a 384-well plate, covered with ultra-high optical clarity film (ThermalSeal RT2RRTM, Z722553, Excel Scientific). The plate was spun down 1,500 g 1 min and incubated for 10 minutes at room temperature. Absorbance for each sample was measured at 540 nm on an Ensight multimode plate reader (PerkinElmer). The readings were then converted to TAG concentrations using standard curves, prepared with triglycerides standard (Randox, 1352TR CAL Standard).

Measurement of transcript levels using qPCR

Several tissue samples containing 3 dissected guts, brains, or heads were collected for each condition or genotype. These samples were then homogenized in 2-ml Eppendorf tubes filled with lysis buffer containing 1% beta-mercaptoethanol, utilizing a TissueLyser LT bead mill (Qiagen) with 5-mm stainless steel beads (Qiagen #69989). RNA extraction was carried out with the NucleoSpin RNA kit (Macherey-Nagel, #740955) following the guidelines provided by the manufacturer. cDNA was synthesized using the High-Capacity cDNA Synthesis kit (Applied Biosystems, #4368814). Quantitative PCR was performed with RealQ Plus 2x Master Mix Green (Ampliqon, #A324402) using a QuantStudio 5 (Applied Biosystems) instrument. Gene expression results were normalized to the housekeeping gene Rp49 using the delta-delta-Ct method. The specific oligonucleotides used are given in Supplementary Table 2.

Statistics

Statistical analyses were performed using the Prism software package (GraphPad, version 10). Data were tested for normality before assessments of significance. For data following a normal distribution, pairwise analyses were conducted using two-tailed unpaired Student’s t-tests, and comparisons involving multiple samples used one-way ANOVA with subsequent post-hoc tests for multiple comparisons. Non-normally distributed data were analyzed using two-tailed unpaired Mann-Whitney U tests or one-way Kruskal-Wallis ANOVA, followed by tests for multiple comparisons. Additionally, interactions between genotype and diet were calculated using 2-way ANOVA. All plots represent the mean ± standard error of the mean (SEM).

eLife Assessment

This important work by Malita et al. describes a mechanism by which an intestinal infection causes an increase in daytime sleep through signaling from the gut to the blood-brain barrier. Their findings suggest that cytokines upd3 and upd2 produced by the intestine following infection act on the glia of the blood-brain barrier to regulate sleep by modulating Allatostatin A signaling. The evidence supporting the claims of the authors is solid. Further verification of certain critical tools, and addressing a few discrepancies from data previously published, would improve this work.

Joint Public Review

Summary:

The authors sought to elucidate the mechanism by which infections increase sleep in Drosophila. Their work is important because it further supports the idea that the blood-brain barrier is involved in brain-body communication, and because it advances the field of sleep research. Using knock-down and knock-out of cytokines and cytokine receptors specifically in the endocrine cells of the gut (cytokines) as well as in the glia forming the blood-brain barrier (BBB) (cytokines receptors), the authors show that cytokines, upd2 and upd3, secreted by entero-endocrine cells in response to infections increase sleep through the Dome receptor in the BBB. They also show that gut-derived Allatostatin (Alst) A promotes wakefulness by inhibiting Alst A signaling that is mediated by Alst receptors expressed in BBB glia. Their results suggest there may be additional mechanisms that promote elevated sleep during gut inflammation.

The authors suggest that upd3 is more critical than upd2, which is not sufficiently addressed or explained. In addition, the study uses the gut's response to reactive oxygen molecules as a proxy for infection, which is not sufficiently justified. Finally, further verification of some fundamental tools used in this paper would further solidify these findings making them more convincing.

Strengths:

(1) The work addresses an important topic and proposes an intriguing mechanism that involves several interconnected tissues. The authors place their research in the appropriate context and reference related work, such as literature about sickness-induced sleep, ROS, the effect of nutritional deprivation on sleep, sleep deprivation and sleep rebound, upregulated receptor expression as a compensatory mechanism in response to low levels of a ligand, and information about Alst A.

(2) The work is, in general, supported by well-performed experiments that use a variety of different tools, including multiple RNAi lines, CRISPR, and mutants, to dissect both signal-sending and receiving sides of the signaling pathway.

(3) The authors provide compelling evidence that shows that endocrine cells from the gut are the source of the upd cytokines that increase daytime sleep, that the glial cells of the BBB are the targets of these upds, and that upd action causes the downregulation of Alst receptors in the BBB via the Jak/Stat pathways.

Weaknesses:

(1) There is a limited characterization of cell types in the midgut which are classically associated with upd cytokine production.

(2) Some of the main tools used in this manuscript to manipulate the gut while not influencing the brain (e.g., Voilà and Voilà + R57C10-GAL80), are not directly shown to not affect gene expression in the brain. This is critical for a manuscript delving into intra-organ communication, as even limited expression in the brain may lead to wrong conclusions.

(3) The model of gut inflammation used by the authors is based on the increase in reactive oxygen species (ROS) obtained by feeding flies food containing 1% H2O2. The use of this model is supported by the authors rather weakly in two papers (refs. 26 and 27 ): The paper by Jiang et al. (ref. 26) shows that the infection by Pseudomonas entomophila induces cytokine responses upd2 and 3, which are also induced by the Jnk pathway. In addition, no mention of ROS could be found in Buchon et al. (ref 27); this is a review that refers to results showing that ROS are produced by the NADPH oxidase DUOX as part of the immune response to pathogens in the gut. Thus, there is no strong support for the use of this model.

(4) Likewise, there is no support for the use of ROS in the food instead a direct infection by pathogenic bacteria. Furthermore, it is known that ROS damages the gut epithelium, which in turn induces the expression of the cytokines studied. Thus the effects observed may not reflect the response to infection. In addition, Majcin Dorcikova et al. (2023). Circadian clock disruption promotes the degeneration of dopaminergic neurons in male Drosophila. Nat Commun. 2023 14(1):5908. doi: 10.1038/s41467-023-41540-y report that the feeding of adult flies with H2O2 results in neurodegeneration if associated with circadian clock defects. Thus, it would be important to discuss or present controls that show that the feeding of H2O2 does not cause neuronal damage.

(5) The novelty of the work is difficult to evaluate because of the numerous publications on sleep in Drosophila. Thus, it would be very helpful to read from the authors how this work is different and novel from other closely related works such as: Li et al. (2023) Gut AstA mediates sleep deprivation-induced energy wasting in Drosophila. Cell Discov. 23;9(1):49. doi: 10.1038/s41421-023-00541-3.

Author response:

The following is the authors’ response to the original reviews.

Joint Public Review:

Summary:

The authors sought to elucidate the mechanism by which infections increase sleep in Drosophila. Their work is important because it further supports the idea that the blood-brain barrier is involved in brain-body communication, and because it advances the field of sleep research. Using knock-down and knock-out of cytokines and cytokine receptors specifically in the endocrine cells of the gut (cytokines) as well as in the glia forming the blood-brain barrier (BBB) (cytokines receptors), the authors show that cytokines, upd2 and upd3, secreted by entero-endocrine cells in response to infections increase sleep through the Dome receptor in the BBB. They also show that gut-derived Allatostatin (Alst) A promotes wakefulness by inhibiting Alst A signaling that is mediated by Alst receptors expressed in BBB glia. Their results suggest there may be additional mechanisms that promote elevated sleep during gut inflammation.

The authors suggest that upd3 is more critical than upd2, which is not sufficiently addressed or explained. In addition, the study uses the gut's response to reactive oxygen molecules as a proxy for infection, which is not sufficiently justified. Finally, further verification of some fundamental tools used in this paper would further solidify these findings making them more convincing.

Strengths:

(1) The work addresses an important topic and proposes an intriguing mechanism that involves several interconnected tissues. The authors place their research in the appropriate context and reference related work, such as literature about sickness-induced sleep, ROS, the effect of nutritional deprivation on sleep, sleep deprivation and sleep rebound, upregulated receptor expression as a compensatory mechanism in response to low levels of a ligand, and information about Alst A.

(2) The work is, in general, supported by well-performed experiments that use a variety of different tools, including multiple RNAi lines, CRISPR, and mutants, to dissect both signal-sending and receiving sides of the signaling pathway.

(3) The authors provide compelling evidence that shows that endocrine cells from the gut are the source of the upd cytokines that increase daytime sleep, that the glial cells of the BBB are the targets of these upds, and that upd action causes the downregulation of Alst receptors in the BBB via the Jak/Stat pathways.

We are pleased that the reviewers recognized the strength and significance of our findings describing a gut-to-brain cytokine signaling mechanism involving the blood-brain barrier (BBB) and its role in regulating sleep, and we thank them for their comments.

Weaknesses:

(1) There is a limited characterization of cell types in the midgut which are classically associated with upd cytokine production.

We thank the reviewer for raising this point. Although several midgut cell types (including the absorptive enterocytes) may indeed produce Unpaired (Upd) cytokines, our study specifically focused on enteroendocrine cells (EECs), which are well-characterized as secretory endocrine cells capable of exerting systemic effects. As detailed in our response to Results point #2 (please see below), we show that EEC-specific manipulation of Upd signaling is both necessary and sufficient to regulate sleep in response to intestinal oxidative stress. These findings support the role of EECs as a primary source of gut-derived cytokine signaling to the brain. To acknowledge the possible involvement of other source, we have also added a statement to the Discussion in the revised manuscript noting that other, non-endocrine gut cell types may contribute to systemic Unpaired signaling that modulates sleep.

(2) Some of the main tools used in this manuscript to manipulate the gut while not influencing the brain (e.g., Voilà and Voilà + R57C10-GAL80), are not directly shown to not affect gene expression in the brain. This is critical for a manuscript delving into intra-organ communication, as even limited expression in the brain may lead to wrong conclusions.

We agree with the reviewer that this is an important point. To address it, we performed additional validation experiments to assess whether the voilà-GAL4 driver in combination with R57C10-GAL80 (EEC>) influences upd2 or upd3 expression in the brain. Our results show that manipulation using EEC> alters upd2 and upd3 expression in the gut (Fig. 1a,b), with new data showing that this does not affect their expression levels in neuronal tissues (Fig. S1a), supporting the specificity of our approach. These new data are now included in the revised manuscript and described in the Results section. This additional validation strengthens our conclusion that the observed sleep phenotypes result from gut-specific cytokine signaling, rather than from effects on Unpaired cytokines produced in the brain.

(1) >(3) The model of gut inflammation used by the authors is based on the increase in reactive oxygen species (ROS) obtained by feeding flies food containing 1% H2O2. The use of this model is supported by the authors rather weakly in two papers (refs. 26 and 27 ): The paper by Jiang et al. (ref. 26) shows that the infection by Pseudomonas entomophila induces cytokine responses upd2 and 3, which are also induced by the Jnk pathway. In addition, no mention of ROS could be found in Buchon et al. (ref 27); this is a review that refers to results showing that ROS are produced by the NADPH oxidase DUOX as part of the immune response to pathogens in the gut. Thus, there is no strong support for the use of this model.

We thank the reviewer for raising this point. We agree that the references originally cited did not sufficiently justify the use of H2O2 feeding as a model of gut inflammation. To address this, we have revised the Results section to clarify that we use H2O2 feeding as a controlled method to elevate intestinal ROS levels, rather than as a general model of inflammation. This approach allows us to investigate the specific effects of ROS-induced cytokine signaling in the gut. We have also added additional citations to support the physiological relevance of this model. For instance, Tamamouna et al. (2021) demonstrated that H2O2 feeding induces intestinal stem-cell proliferation – a response also observed during bacterial infection – and Jiang et al. (2009) showed that enteric infections increase upd2 and upd3 expression, which we similarly observe following H2O2 feeding (Fig. 3a). These findings support the use of H2O2 as a tool to mimic specific ROS-linked responses in the gut. We believe this targeted and tractable model is a strength of our study, enabling us to dissect how intestinal ROS modulates systemic physiology through cytokine signaling

Additionally, we have included a statement in the Discussion acknowledging that ROS generated during infection may activate signaling mechanisms distinct from those triggered by chemically induced oxidative stress, and that exploring these differences in future studies may yield important insights into gut–brain communication. These revisions provide a stronger justification for our model while more accurately conveying both its relevance and its limitations.

(2) >(4) Likewise, there is no support for the use of ROS in the food instead a direct infection by pathogenic bacteria. Furthermore, it is known that ROS damages the gut epithelium, which in turn induces the expression of the cytokines studied. Thus the effects observed may not reflect the response to infection. In addition, Majcin Dorcikova et al. (2023). Circadian clock disruption promotes the degeneration of dopaminergic neurons in male Drosophila. Nat Commun. 2023 14(1):5908. doi: 10.1038/s41467-02341540-y report that the feeding of adult flies with H2O2 results in neurodegeneration if associated with circadian clock defects. Thus, it would be important to discuss or present controls that show that the feeding of H2O2 does not cause neuronal damage.

We thank the reviewer for this thoughtful follow-up point. We would like to clarify that we do not claim that the effects observed in our study directly reflect the full response to enteric infection. As outlined in our revised response to comment 3, we have updated the manuscript to more precisely describe the H2O2-feeding paradigm as a model that induces local intestinal ROS responses comparable to, but not equivalent to, those observed during pathogenic challenges. This revised framing highlights both the potential similarities and differences between chemically induced oxidative stress and infection-induced responses. Indeed, in the revised Discussion, we now explicitly acknowledge that ROS generated during infection may engage distinct signaling mechanisms compared to exogenous H2O2 and emphasize the value of future studies in delineating these pathways. We are currently pursuing this direction in an independent ongoing study investigating the effects of enteric infections. However, for the present work, we chose to focus on the effects of ROS-induced responses in isolation, as this provides a clean and well-controlled context to dissect the specific contribution of oxidative stress to cytokine signaling and sleep regulation.

To further address the reviewer’s concern, we have also included new data (a TUNEL stain for apoptotic DNA fragmentation) in the revised manuscript showing that H2O2 feeding does not damage neuronal tissues under our experimental conditions (Fig. S3f,g). This addresses the point raised regarding the potential neurotoxicity of H2O2, as described by Majcin Dorcikova et al. (2023), and supports the specificity of the sleep phenotypes observed in our study. We believe these revisions and clarifications strengthen the manuscript and make our interpretation more precise.

(3) >(5) The novelty of the work is difficult to evaluate because of the numerous publications on sleep in Drosophila. Thus, it would be very helpful to read from the authors how this work is different and novel from other closely related works such as: Li et al. (2023) Gut AstA mediates sleep deprivation-induced energy wasting in Drosophila. Cell Discov. 23;9(1):49. doi: 10.1038/s41421-023-00541-3.

Our work highlights a distinct role for gut-derived AstA in sleep regulation compared to findings by Lin et al. (Cell Discovery, 2023)[1], who showed that gut AstA mediates energy wasting during sleep deprivation. Their study focused on the metabolic consequences of sleep loss, proposing that sleep deprivation increases ROS in the gut, which then promotes the release of the glucagon-like hormone adipokinetic hormone (AKH) through gut AstA signaling, thereby triggering energy expenditure.

In contrast, our study addresses the inverse question – how ROS in the gut influences sleep. In our model, intestinal ROS promotes sleep, raising the intriguing possibility – cleverly pointed out by the reviewers – that ROS generated during sleep deprivation might promote sleep by inducing Unpaired cytokine signaling in the gut. According to our findings, this suppresses wake-promoting AstA signaling in the BBB, providing a mechanism to promote sleep as a restorative response to gut-derived oxidative stress and potentially limiting further ROS accumulation. Importantly, our findings support a wakepromoting role for EEC-derived AstA, demonstrated by several lines of evidence. First, EEC-specific knockdown of AstA increases sleep. Second, activation of AstA+ EECs using the heat-sensitive cation channel Transient Receptor Potential A1 (TrpA1) reduces sleep, and this effect is abolished by simultaneous knockdown of AstA, indicating that the sleep-suppressing effect is mediated by AstA and not by other peptides or secreted factors released by these cells. Third, downregulation of AstA receptor expression in BBB glial cells increases sleep, further supporting the existence of a functional gut AstA– glia arousal pathway. We have now included new data in the revised manuscript showing that AstA release from EECs is downregulated during intestinal oxidative stress (Fig. 7k,l,m). This suggests that this wake-promoting signal is suppressed both at its source (the gut endocrine cells), by unknown means, and at its target, the BBB, via Unpaired cytokine signaling that downregulates AstA receptor expression. This coordinated downregulation may serve to efficiently silence this arousal-promoting pathway and facilitate sleep during intestinal stress. These new data, along with an expanded discussion, provide further mechanistic insight into gut-derived AstA signaling and strengthen our proposed model.

This contrasts with the interpretation by Lin et al., who observed increased AstA peptide levels in EECs after antioxidant treatment and interpreted this as peptide retention. However, peptide accumulation may result from either increased production or decreased release, and peptide levels alone are insufficient to distinguish between these possibilities. To resolve this, we examined AstA transcript levels, which can serve as a proxy for production. Following oxidative stress (24 h of 1% H2O2 feeding and the following day), when animals show increased sleep (Fig. 7e), we observed a decrease in AstA transcript levels followed by an increase in peptide levels (Fig. 7k,l,m), suggesting that oxidative stress leads to reduced gut AstA production and release. Furthermore, we recently found that a class of EECs that produce the hormone Tachykinin (Tk) and are distinct from the AstA+ EECs express the ROSsensitive cation channel TrpA1 (Ahrentløv et al., 2025, Nature Metabolism2). In these Tk+ EECs, TrpA1 mediates ROS-induced Tk hormone release. In contrast, single-cell RNA-seq data[3] do not support TrpA1 expression in AstA+ EECs, consistent with our findings that ROS does not promote AstA release – an effect that would be expected if TrpA1 were functionally expressed in AstA+ EECs. This contradicts the findings of Lin et al., who reported TrpA1 expression in AstA+ EECs. We have now included relevant single-cell data in the revised manuscript (Fig. S6f) showing that TrpA1 is specifically expressed in Tk+ EECs, but not in AstA+ EECs, and we have expanded the discussion to address discrepancies in TrpA1 expression and AstA regulation.

Taken together, our results reveal a dual-site regulatory mechanism in which Unpaired cytokines released from the gut act at the BBB to downregulate AstA receptor expression, while AstA release from EECs is simultaneously suppressed. We thank the reviewers for raising this important point. We have also included a discussion the other point raised by the reviewers – the possibility that ROS generated during sleep deprivation may engage the same signaling pathways described here, providing a mechanistic link between sleep deprivation, intestinal stress, and sleep regulation.

Recommendations for the authors:

A- Material and Methods:

(1) Feeding Assay: The cited publication (doi.org:10.1371/journal.pone.0006063) states: "For the amount of label in the fly to reflect feeding, measurements must therefore be confined to the time period before label egestion commences, about 40 minutes in Drosophila, a time period during which disturbance of the flies affects their feeding behavior. There is thus a requirement for a method of measuring feeding in undisturbed conditions." Was blue fecal matter already present on the tube when flies were homogenized at 1 hour? If so, the assay may reflect gut capacity rather than food passage (as a proxy for food intake). In addition, was the variability of food intake among flies in the same tube tested (to make sure that 1-2 flies are a good proxy for the whole population)?

We agree that this is an important point for feeding experiments. We are aware of the methodological considerations highlighted in the cited study and have extensive experience using a range of feeding assays in Drosophila, including both short- and long-term consumption assays (e.g., dye-based and CAFE assays), as well as automated platforms such as FLIC and FlyPAD (Nature Communications, 2022; Nature Metabolism, 2022; and Nature Metabolism, 2025)[2,4,5].

For the dye-based assay, we carefully selected a 1-hour feeding window based on prior optimization. Since animals were not starved prior to the assay, shorter time points (e.g., 30 minutes) typically result in insufficient ingestion for reliable quantification. A 1-hour period provides a robust readout while remaining within the timeframe before significant label excretion occurs under our experimental conditions. To support the robustness of our findings, we complemented the dye-based assay with data from FLIC, which enables automated, high-resolution monitoring of feeding behavior in undisturbed animals over extended periods. The FLIC results were consistent with the dye-based data, strengthening our confidence in the conclusions. To minimize variability and ensure consistency across experiments, all feeding assays were performed at the same circadian time – Zeitgeber Time 0 (ZT0), corresponding to 10:00 AM when lights are turned on in our incubators. This time point coincides with the animals' natural morning feeding peak, allowing for reproducible comparisons across conditions. Regarding variability among flies within tubes, each biological replicate in the dye assay consisted of 1–2 flies, and results were averaged across multiple replicates. We observed good consistency across samples, suggesting that these small groups reliably reflect group-level feeding behavior under our conditions.

(2) Biological replicates: whereas the number of samples is clearly reported in each figure, the number of biological replicates is not indicated. Please include this information either in Material and methods or in the relevant figure legends. Please also include a description of what was considered a biological replicate.

We have now clarified in the Materials and Methods section under Statistics that all replicates represent independent biological samples, as suggested by the reviewers.

(3) Control Lines: please indicate which control lines were used instead of citing another publication. If preferred, this information could be supplied as a supplementary table.

We now provide a clear description of the control lines used in the Materials and Methods section. Specifically, all GAL4 and GAL80 lines used in this study were backcrossed for several generations into a shared w1118 background and then crossed to the same w1118 strain used as the genetic background for the UAS-RNAi, <i.CRISPR, or overexpression lines. This approach ensures, to a strong approximation, that the only difference between control and experimental animals is the presence or absence of the UAS transgene.

(4) Statistical analyses: for some results (e.g., those shown in Figure 3d), it could be useful to test the interaction between genotype and treatment.

We thank the reviewer for this helpful suggestion. In response, we have now performed two-way ANOVA analyses to assess genotype × treatment (diet) interaction effects for the relevant data, including those shown in Figure 3d as well as additional panels where animals were exposed to oxidative stress and sleep phenotypes were measured. We have added the corresponding interaction p-values in the updated figure legends for Figures 3d, 3k, 5a–c, 5f, 5h, 5i, 6c, 6e, and 7e. All of these tests revealed significant interaction effects, supporting the conclusion that the observed differences in sleep phenotypes are specifically dependent on the interaction between genetic manipulation (e.g., cytokine or receptor knockdown) and oxidative stress. These additions reinforce the interpretation that Unpaired cytokine signaling, glial JAK-STAT pathway activity, and AstA receptor regulation functionally interact with intestinal ROS exposure to modulate sleep. We thank the reviewer for suggesting this improvement.

(5) Reporting of p values. Some are reported as specific values whereas others are reported as less than a specific value. Please make this reporting consistent across different figures.

All p-values reported in the manuscript are exact, except in cases where values fall below p < 0.0001. In those instances, we use the inequality because the Prism software package (GraphPad, version 10), which was used for all statistical analyses, does not report more precise values. We believe this reporting approach reflects standard practice in the field.

(6) Please include the color code used in each figure, either in the figure itself or in the legend.

We have now clarified the color coding in all relevant figures. In particular, we acknowledge that the meaning of the half-colored circles used to indicate H2O2 treatment was not previously explained. These have now been clearly labeled in each figure to indicate treatment conditions.

(7) The scheme describing the experimental conditions and the associated chart is confusing. Please improve.

We have improved the schematic by replacing “ROS” with “H2O2” to more clearly indicate the experimental condition used. Additionally, we have added the corresponding circle annotations so that they now also appear consistently above the relevant charts. This revised layout enhances clarity and helps readers more easily interpret the experimental conditions. We believe these changes address the reviewer’s concern and make the figure significantly more intuitive.

1. Please indicate which line was used for upd-Gal4 and the evidence that it faithfully reflects upd3 expression.

We have now clarified in the Materials and Methods section that the upd3-GAL4 line used in our study is Bloomington stock #98420, which drives GAL4 expression under the control of approximately 2 kb of sequence upstream of the upd3 start codon. This line has previously been used as a transcriptional reporter for upd3 activity. The only use of this line was to illustrate reporter expression in the EECs. To support this aspect of Upd3 expression, we now include new data in the revised manuscript using fluorescent in situ hybridization (FISH) against upd3, which confirms the presence of upd3 transcripts in prospero-positive EECs of the adult midgut (Fig. S1b). Additionally, we show that upd3 transcript levels are significantly reduced in dissected midguts following EEC-specific knockdown using multiple independent RNAi lines driven by voilà-GAL4, both alone and in combination with R57C10-GAL80, consistent with endogenous expression in these cells (Fig. 1a,b).

To further address the reviewer’s concern and provide additional support for the endogenous expression of upd3 in EECs, we performed targeted knockdown experiments focusing on molecularly defined EEC subpopulations. The adult Drosophila midgut contains two major EEC subtypes characterized by their expression of Allatostatin C (AstC) or Tachykinin (Tk), which together encompass the vast majority of EECs. To selectively manipulate these populations, we used AstC-GAL4 and Tk-GAL4 drivers – both knock-in lines in which GAL4 is inserted at the respective endogenous hormone loci. This design enables precise GAL4 expression in AstC- or Tk-expressing EECs based on their native transcriptional profile. To eliminate confounding neuronal expression, we combined these drivers with R57C10GAL80, restricting GAL4 activity to the gut and generating AstCGut> and TkGut> drivers. Using these tools, we knocked down upd2 and upd3 selectively in the AstC- or Tk-positive EECs. Knockdown of either cytokine in AstC-positive EECs significantly increased sleep under homeostatic conditions, recapitulating the phenotype observed with knockdown in all EECs (Fig. 1m-o). In contrast, knockdown of upd2 or upd3 in Tk-positive EECs had no effect on sleep (Fig. 1p-r). Furthermore, we show in the revised manuscript that selective knockdown of upd2 or upd3 in AstC-positive EECs abolishes the H2O2-induced increase in sleep (Fig. 3f–h). These findings demonstrate that Unpaired cytokine signaling from AstC-positive EECs is essential for mediating the sleep response to intestinal oxidative stress, highlighting this specific EEC subtype as a key source of cytokine-driven regulation in this context. These new results indicate that AstC-positive EECs are a primary source of the Unpaired cytokines that regulate sleep, while Tk-positive EECs do not appear to contribute to this function. Importantly, upd3 transcript levels were significantly reduced in dissected midguts following AstCGut driven knockdown (Fig. S1r), further confirming that upd3 is endogenously expressed in AstC-positive EECs. Thus we have bolstered our confidence that upd3 is indeed expressed in EECs, as illustrated by the reporter line, through several means.

(9) Please indicate which GFP line was used with upd-Gal4 (CD8, NLS, un-tagged, etc). The Material and Methods section states that it was "UAS-mCD8::GFP (#5137);", however, the stain does not seem to match a cell membrane pattern but rather a nuclear or cytoplasmic pattern. This information would help the interpretation of Figure 1C.

We confirm that the GFP reporter line used with upd3-GAL4 was obtained from Bloomington stock #98420. As noted by the Bloomington Drosophila Stock Center, “the identity of the UAS-GFP transgene is a guess,” and the subcellular localization of the GFP fusion is therefore uncertain. We agree with the reviewer that the signal observed in Figure 1c does not display clear membrane localization and instead appears diffuse, consistent with cytoplasmic or partially nuclear localization. In any case, what we find most salient is the reporter’s labeling of Prospero-positive EECs in the adult midgut, consistent with upd3 expression in these cells. This conclusion is further supported by multiple lines of evidence presented in the revised manuscript, as mentioned above in response to question #8: (1) fluorescent in situ hybridization (FISH) for upd3 confirms expression in EECs (Fig. S1b), (2) EEC-specific RNAi knockdown of upd3 reduces transcript levels in dissected midguts, and (3) publicly available single-cell RNA sequencing datasets[3] also indicate that upd3 is expressed at low levels in a subset of adult midgut EECs under normal conditions. We have also clarified in the revised Materials and Methods section that GFP localization is undefined in the upd3-GAL4 line, to guide interpretation of the reporter signal.

B- Results

(1) Figure 1: According to previous work (10.1016/j.celrep.2015.06.009, http://flygutseq.buchonlab.com/data?gene=upd3%0D%0A), in basal conditions upd3 is expressed as following: ISC (35 RPKM), EB (98 RPKM), EC (57 RPKM), and EEC (8 RPKM). Accordingly, even complete KO in EECs should eliminate only a small fraction of upd3 from whole guts, even less considering the greater abundance of other cell types such as ECs compared to EECs. It would be useful to understand where this discrepancy comes from, in case it is affecting the conclusion of the manuscript. While this point per se does not affect the main conclusions of the manuscript, it makes the interpretation of the results more difficult.

We acknowledge the previously reported low expression of upd3 in EECs. However, the FlyGut-seq site appears to be no longer available, so we could not directly compare other related genes. Nonetheless, our data – based on in situ hybridization, reporter expression, and multiple RNAi knockdowns – consistently support upd3 expression in EECs. These complementary approaches strengthen the conclusion that EECs are an important source of systemic upd3 under the conditions tested.

(2) Figure 1: The upd2-3 mutants show sleep defects very similar to those of EEC>RNAi and >Cas9. It would thus be helpful to try to KO upd3 with other midgut drivers (An EC driver like Myo1A or 5966GS and a progenitor driver like Esg or 5961GS) to validate these results. Such experiments might identify precisely which cells are involved in the gut-brain signaling reported here.

We appreciate the reviewer’s suggestion and agree that exploring other potential sources of Upd3 in the gut is an interesting direction. In this study, we have focused on EECs, which are the primary hormone-secreting cells in the intestine and thus the most likely candidates for mediating systemic effects such as gut-to-brain signaling. While it is possible that other gut cell types – such as enterocytes (e.g., Myo1A+) or intestinal progenitors (e.g., Esg+) – also contribute to Upd3 production, these cells are not typically endocrine in nature. Demonstrating their involvement in gutto-brain communication would therefore require additional, extensive validation beyond the scope of the current study. Importantly, our data show that manipulating Upd3 specifically in EECs is both necessary and sufficient to modulate sleep in response to intestinal ROS, strongly supporting the conclusion that EEC-derived cytokine signaling underlies the observed phenotype. In contrast, manipulating cytokines in other gut cells could produce indirect effects – such as altered proliferation, epithelial integrity, or immune responses – that complicate the interpretation of behavioral outcomes like sleep. For these reasons, we chose to focus on EECs as the source of endocrine signals mediating gut-to-brain communication. However, to address this point raised by the reviewer, we have now included a statement in the Discussion acknowledging that other non-endocrine gut cell types may also contribute to the systemic Unpaired signaling that modulates sleep in response to intestinal oxidative stress.

(3) Figure 3: "This effect mirrored the upregulation observed with EEC-specific overexpression of upd3, indicating that it reflects physiologically relevant production of upd3 by the gut in response to oxidative stress." Please add (Figure 3a) at the end of this sentence.

We have now added “(Figure 3a)” at the end of the sentence to clearly reference the relevant data.

(4) For Figure 3b, do you have data showing that the increased amount of sleep was due to the addition of H2O2 per se, rather than the procedure of adding it?

We have added new data to address this point. To ensure that the observed sleep increase was specifically due to the presence of H2O2 and not an effect of the food replacement procedure, we performed a control experiment in which animals were fed standard food prepared using the same protocol and replaced daily, but without H2O2. These animals did not exhibit increased sleep, confirming that the sleep effect is attributable to intestinal ROS rather than the supplementation procedure itself (Fig. S3a). Thanks for the suggestion.

(5) In the text it is stated that "Since 1% H2O2 feeding induced robust responses both in upd3 expression and in sleep behavior, we asked whether gut-derived Unpaired signaling might be essential for the observed ROS-induced sleep modulation. Indeed, EEC-specific RNAi targeting upd2 or upd3 abolished the sleep response to 1% H2O2 feeding." While it is indeed true that there is no additional increase in sleep time due to EEC>upd3 RNAi, it is also true that EEC>upd3 RNAi flies, without any treatment, have already increased their sleep in the first place. It is then possible that rather than unpaired signaling being essential, an upper threshold for maximum sleep allowed by manipulation of these processes was reached. It would be useful to discuss this point.

Several findings argue against a ceiling effect and instead support a requirement for Unpaired signaling in mediating ROS-induced sleep. Animals with EEC-specific upd2 or upd3 knockdown or null mutation not only fail to increase sleep following H2O2 treatment but actually exhibit reduced sleep during oxidative stress (Fig. 3e, k, l; Fig. 5e, f), suggesting that Unpaired signaling is required to sustain sleep under these conditions. Similarly, animals with glial dome knockdown also show reduced sleep under oxidative stress, closely mirroring the phenotype of EEC-specific upd3 RNAi animals (Fig. 5a–c, g–i). These results support the conclusion that gut-to-glia Unpaired cytokine signaling is necessary for maintaining elevated sleep during oxidative stress. In the absence of this signaling, animals exhibit increased wakefulness. We identify AstA as one such wake-promoting signal that is suppressed during intestinal stress. We present new data showing that this pathway is downregulated not only via Unpaired-JAK/STAT signaling in glial cells but also through reduced AstA release from the gut in the revised manuscript. This model, in which Unpaired cytokines promote sleep during intestinal stress by suppressing arousal pathways, is discussed throughout the manuscript to address the reviewer’s point.

(6) In Figure 3k, the dots highlighting the experiment show an empty profile, a full one, and a half one. Please define what the half dots represent.

We have now clarified the color coding in all relevant figures. Specifically, we acknowledge that the meaning of the half-colored circles indicating H2O2 treatment was not previously defined – it indicates washout or recovery time. In the revised version, these symbols are now clearly labeled in each figure to indicate the treatment condition, ensuring consistent and intuitive interpretation across all panels.

(7) The authors used appropriate GAL4 and RNAi lines to the knockdown dome, a upd2/3 JAK-STATlinked receptor, specifically in neurons and glia, respectively, in order to identify the CNS targets of upd2/3 cytokines produced by enteroendocrine cells (EECs). Pan-neuronal dome knockdown did not alter daytime sleep in adult females, yet pan-glial dome knockdown phenocopied effects of upd2/3 knockdown in EECs. They also observed that EEC-specific knockdown of upd2 and upd3 led to a decrease in JAK-STAT reporter activity in repo-positive glial cells. This supports the authors' conclusion that glial cells, not neurons, are the targets by which unpaired cytokines regulate sleep via JAK-STAT signaling. However, they do not show nighttime sleep data of pan-neuronal and pan-glial dome knockdowns. It would strengthen their conclusion if the nighttime sleep of pan-glial dome knockdown phenocopied the upd2/3 knockdowns as well, provided the pan-neuronal dome knockdown did not alter nighttime sleep.

We have now added nighttime sleep data for both pan-glial and pan-neuronal domeless knockdowns in the revised manuscript (Fig. 2a). Glial knockdown increased nighttime sleep, similar to EEC-specific upd2/3 knockdown, while neuronal knockdown had no effect. These results further support the glial cells’ being the relevant target of gut-derived Unpaired signaling.

(8) The authors only used one method to induce oxidative stress (hydrogen peroxide feeding). It would strengthen their argument to test multiple methods of inducing oxidative stress, such as lipopolysaccharide (LPS) feeding. In addition, it would be useful to use a direct bacterial infection to confirm that in flies, the infection promotes sleep. Additionally, flies deficient in Dome in the BBB and infected should not be affected in their sleep by the infection. These experiments would provide direct support for the mechanism proposed. Finally, the authors should add a primary reference for using ROS as a model of bacterial infection and justify their choice better.

We agree that directly comparing different models of intestinal stress, such as bacterial infection or LPS feeding, would provide valuable insight into how gut-derived signals influence sleep in response to infection. As noted in our detailed responses above, we now include an expanded rationale for our use of H2O2 feeding as a controlled and well-established method for inducing intestinal ROS – one of the key physiological responses to enteric infection and inflammation. In the revised Discussion, we explicitly acknowledge that pathogenic infections – which trigger both intestinal ROS and additional immune pathways – may engage distinct or complementary mechanisms compared to chemically induced oxidative stress. We emphasize the importance of future studies aimed at dissecting these differences. In fact, we are actively pursuing this direction in ongoing work examining sleep responses to enteric infection. For the purposes of the present study, however, we chose to focus on a tractable and specific model of ROS-induced stress to define the contribution of Unpaired cytokine signaling to gut-brain communication and sleep regulation. This approach allowed us to isolate the effect of oxidative stress from other confounding immune stimuli and identify a glia-mediated signaling mechanism linking gut epithelial stress to changes in sleep behavior.

(9) To confirm that animals lacking EEC Unpaired signaling are not more susceptible to ROS-induced damage, the authors assessed the survival of upd2 and upd3 knockdowns on 1% H2O2 and concluded they display no additional sensitivity to oxidative stress compared to controls. It may be useful to include other tests of sensitivity to oxidative stress, in addition to survival.

We appreciate the reviewer’s suggestion. In our view, survival is a highly informative and stringent readout, as it reflects the overall physiological capacity of the animal to withstand oxidative stress. Importantly, our data show that animals lacking EEC-derived Unpaired signaling do not exhibit reduced survival following H2O2 exposure, indicating that their oxidative stress resistance is not compromised. Furthermore, we previously confirmed that feeding behavior is unaffected in these animals, suggesting that their ability to ingest food (and thus the stressor) is not impaired. As a molecular complement to these assays in response to this point and others, we have also performed an assessment of neuronal apoptosis (a TUNEL assay, Fig. S3f,g). This assay did not identify an increase in cell death in the brains of animals fed peroxide-containing medium. Thus, gross neurological health, behavior, and overall survival appear to be resilient to the environmental treatment regime we apply here, suggesting that the outcomes we observe arise from signaling per se.

(10) The authors confirmed that animals lacking EEC-derived upd3 displayed sleep suppression similar to controls in response to starvation. These results led the authors to conclude that there is a specific requirement for EEC-derived Unpaired signaling in responding to intestinal oxidative stress. However, they previously showed that EEC-specific knockdown of upd3 and upd2 led to increased daytime sleep under normal feeding conditions. Their interpretations of their data are inconsistent.

We appreciate the reviewer’s comment. While animals lacking EEC-derived Unpaired signaling show increased baseline sleep under normal feeding conditions, they still exhibit a robust reduction in sleep when subjected to starvation – comparable to that of control animals (Fig. S3h–j). This demonstrates that they retain the capacity to appropriately modulate sleep in response to metabolic stress. Thus, the sleep-promoting phenotype under normal conditions does not reflect a generalized inability to adjust sleep behavior. Rather, it highlights a specific role for Unpaired signaling in mediating sleep responses to intestinal oxidative stress, not in broadly regulating all sleep-modulating stimuli.

(11) The authors report a significant increase in JAK-STAT activity in surface glial cells at ZT0 in animals fed 1% H2O2-containing food for 20 hours. This response was abolished in animals with EECspecific knockdown of upd2 or upd3. The authors confirmed there were no unintended neuronal effects on upd2 or upd3 expression in the heads. They also observed an upregulation of dome transcript levels in the heads of animals with EEC-specific knockdown of upd3 fed 1% H2O2-containing food for 15 hours, which they interpret to be a compensatory mechanism in response to low levels of the ligand. This assay is inconsistent with previous experiments in which animals were fed hydrogen peroxide for 20 hours.

We thank the reviewer for identifying this discrepancy. The inconsistency arose from a labeling error in the manuscript. Both the JAK-STAT reporter assays in glial cells and the dome expression measurements were performed following 15 hours of H2O2 feeding, not 20 hours as previously stated. We have now corrected this in the revised manuscript.

(12) The authors show that animals with glia-specific dome knockdown did not have decreased survival on H2O2-containing food, and displayed normal rebound sleep in the morning following sleep deprivation. These results potentially undermine the significance of the paper. If the normal sleep response to oxidative stress is an important protective mechanism, why would oxidative stress not decrease survival in dome knockdown flies (that don't have the normal sleep response to oxidative stress)? This suggests that the proposed mechanism is not important for survival. The authors conclude that Dome-mediated JAK-STAT signaling in the glial cells specifically regulates ROS-induced sleep responses, which their results support.

We agree that our survival data show that glial dome knockdown does not reduce survival under continuous oxidative stress. However, we believe this does not undermine the importance of the sleep response as an adaptive mechanism. In our survival assay, animals were continuously exposed to 1% H2O2 without the opportunity to recover. In contrast, under natural conditions, oxidative stress is likely to be intermittent, and the ability to mount a sleep response may be particularly important for promoting recovery and maintaining homeostasis during or after transient stress episodes. Thus, while the JAK-STAT-mediated sleep response may not directly enhance survival under constant oxidative challenge, it likely plays a critical role in adaptive recovery under natural conditions.

(13) Altogether, the authors conclude that enteric oxidative stress induces the release of Unpaired cytokines which activate the JAK-STAT pathway in subperineurial glia of the BBB, which leads to the glial downregulation of receptors for AstA, which is a wake-promoting factor also released by EECs. This mechanism is supported by their results, however, this research raises some intriguing questions, such as the role of upd2 versus upd3, the role of AstA-R1 versus AstA-R2, the importance of this mechanism in terms of survival, the sex-specific nature of this mechanism, and the role that nutritional availability plays in the dual functionality of Unpaired cytokine signaling in regards to sleep.

We thank the reviewer for highlighting these important questions. Our data suggest that Upd2 and Upd3, while often considered partially redundant, both contribute to sleep regulation, with stronger effects observed for Upd3. This is consistent with prior studies indicating overlapping but non-identical roles for these cytokines. Similarly, although AstA-R1 and AstA-R2 can both be activated by AstA, knockdown of AstA-R2 consistently produces more robust sleep phenotypes, suggesting a predominant role in mediating this effect. The possibility of sex-specific regulation is indeed compelling. While our study focused on females, many gut hormones show sex-dependent activity, and we recognize this as an important avenue for future research. Finally, we have included new data in the revised manuscript showing that gut-derived AstA is downregulated under oxidative stress, further supporting our model in which Unpaired signaling suppresses arousal pathways during intestinal stress

(14)Data Availability: It is indicated that: "Reasonable data requests will be fulfilled by the lead author". However, eLife's guidelines for data sharing require that all data associated with an article to be made freely and widely available.

We thank the reviewer for pointing this out. We have revised the Data Availability section of the manuscript to clarify that all data will be made freely available from the lead contact without restriction, in accordance with eLife’s open data policy.

References

(1) Li, Y., Zhou, X., Cheng, C., Ding, G., Zhao, P., Tan, K., Chen, L., Perrimon, N., Veenstra, J.A., Zhang, L., and Song, W. (2023). Gut AstA mediates sleep deprivaPon-induced energy wasPng in Drosophila. Cell Discov 9, 49. 10.1038/s41421-023-00541-3. (2) Ahrentlov, N., Kubrak, O., Lassen, M., Malita, A., Koyama, T., Frederiksen, A.S., Sigvardsen, C.M., John, A., Madsen, P., Halberg, K.A., et al. (2025). Protein-responsive gut hormone Tachykinin directs food choice and impacts lifespan. Nature Metabolism. 10.1038/s42255-025-01267-0.

(3) Li, H., Janssens, J., De Waegeneer, M., Kolluru, S.S., Davie, K., Gardeux, V., Saelens, W., David, F.P.A., Brbic, M., Spanier, K., et al. (2022). Fly Cell Atlas: A single-nucleus transcriptomic atlas of the adult fruit fly. Science 375, eabk2432. 10.1126/science.abk2432.

(4) Kubrak, O., Koyama, T., Ahrentlov, N., Jensen, L., Malita, A., Naseem, M.T., Lassen, M., Nagy, S., Texada, M.J., Halberg, K.V., and Rewitz, K. (2022). The gut hormone AllatostaPn C/SomatostaPn regulates food intake and metabolic homeostasis under nutrient stress. Nature communicaPons 13, 692. 10.1038/s41467-022-28268-x.

(5) Malita, A., Kubrak, O., Koyama, T., Ahrentlov, N., Texada, M.J., Nagy, S., Halberg, K.V., and Rewitz, K. (2022). A gut-derived hormone suppresses sugar appePte and regulates food choice in Drosophila. Nature Metabolism 4, 1532-1550. 10.1038/s42255-022-00672-z.

Version 2

Abstract

Abstract

Sickness-induced sleep is a behavior conserved across species that promotes recovery from illness, yet the underlying mechanisms are poorly understood. Here, we show that interleukin-6-like cytokine signaling from the Drosophila gut to brain glial cells regulates sleep. Under healthy conditions, this pathway promotes wakefulness. However, elevated gut cytokine signaling in response to oxidative stress – triggered by immune and inflammatory responses in the intestine – induces sleep. The cytokines Unpaired 2 and -3 are upregulated by oxidative stress in enteroendocrine cells and activate JAK-STAT signaling in glial cells, including those of the blood-brain barrier (BBB). This activity maintains elevated sleep during oxidative-stress-induced intestinal disturbances, suggesting that the JAK-STAT pathway in glia inhibits wake-promoting signaling to facilitate sleep-dependent restoration under these conditions. We find that the enteric peptide Allatostatin A (AstA) enhances wakefulness, and during intestinal oxidative stress, gut-derived Unpaired 2/3 inhibits AstA receptor expression in BBB glia, thereby sustaining an elevated sleep state during gut inflammation or illness. Taken together, our work identifies a gut-to-glial communication pathway that couples sleep with intestinal homeostasis and disease, enhancing sleep during intestinal sickness, and contributes to our understanding of how sleep disturbances arise from gastrointestinal disturbances.

Introduction

Sleep is a conserved behavior essential for physical health and mental well-being. This process maintains physiological balance and promotes recovery from illnesses and other stressors1,2. In healthy states, animals exhibit rhythmic periods of wakefulness and activity, but their sleep patterns significantly increase during illness. This adaptive behavior is related to the fundamental role of sleep in the recovery process, allowing the body to conserve energy and allocate resources towards eliminating pathogens and repairing tissue damage. Sleep patterns are generated by neural circuits within the brain. These circuits engage in complex interactions involving diverse brain regions, neurotransmitters, and signaling pathways to regulate the cycles of sleep and wakefulness3,4. For sleep to be effectively modulated during illness, there must be a dynamic interaction between the physiological states of the body’s organs and these central sleep-regulatory circuits. However, the signals that mediate this communication and the mechanisms by which they modulate sleep during health and disease remain poorly defined.

Sickness-induced sleep, a behavior conserved across species including mammals and flies5,6, is influenced by cytokines, which are key mediators of immune and inflammatory responses1,2. Cytokines such as Interleukin 1 (IL-1) and tumor necrosis factor alpha (TNFα) are expressed in the healthy mammalian brain in regions that are implicated in sleep regulation, and their circulating levels change during the normal sleep-wake cycle, peaking during the sleep phase. Furthermore, these factors’ effects on sleep appear to be dose-dependent, as low levels of IL-1 can enhance sleep, whereas higher doses can inhibit sleep, indicating a dual functionality. Since immune responses alter the expression of these cytokines, they have been hypothesized to act as “somnogens” that promote sleep during times of infection or illness. However, the connection between sleep and immune function is bidirectional, since sleep deprivation in mammals has been linked to increased inflammatory response via IL-1 and TNFα2. In the fruit fly Drosophila, sleep deprivation also seems to influence TNFα in astrocyte-like cells to regulate homeostatic sleep responses that enable sleep rebound after deprivation7. However, the effect of these cytokines on sleep has mostly been linked to their central expression and function within the central nervous system (CNS), while the coupling of cytokines produced by peripheral tissues to sleep-regulatory systems within the brain remains poorly understood.

Disorders affecting the gastrointestinal tract can lead to sleep disturbances8, which are also associated with virtually all mental illnesses9-11. Conditions including depression, anxiety, and disturbed sleep are frequently observed in individuals with gut inflammation, and the gut microbiome has also been linked to sleep quality and mental health8,12-14. These associations suggest a strong connection between gut health and sleep. For changes in gut status to bring about behavioral changes, the gut must sense its state of health, damage, or presence of pathogens and release signals that lead to altered cellular function within the brain. This gut-to-brain signaling is mediated in large part by hormonal factors released from specialized endocrine cells of the gut, the enteroendocrine cells (EECs)15,16. Like the mammalian intestine, the Drosophila gut produces numerous diverse hormones from specialized EECs17-22. Some of these gut hormones are released in response to nutritional intake, and they diet-dependently modulate sleep patterns and arousability through communication with neuroendocrine centers and brain circuits23-25.

In the fly, enteric infection or damage leads to the production of reactive oxygen species (ROS) and the increased expression of the IL-6-like inflammatory cytokines Unpaired 2 and -3 (Upd2/3) in the absorptive enterocytes, a response required for local gut regeneration26,27. These cytokine factors signal through their receptor, Domeless (Dome), to activate the JAK/STAT signaling pathway in target cells, which is important for both immune function and metabolism in flies, demonstrating a conserved function of cytokine action in this species. While the three related cytokines Upd1, Upd2, and Upd3 all signal through Dome, Upd2 and in particular Upd3 are IL-6-like cytokines mainly triggered by infection and are directly linked with cellular immune responses28-30. As in mammals, cytokines are also produced centrally within the Drosophila brain, and neuronal Upd1 acts in a leptin-like manner to regulate feeding31, a behavior that is also linked with sleep4. Peripheral cytokine signaling has also been shown to modulate sleep in this species, where the fat tissue releases Upd2 to reflect adequate nutrition, and this signal modulates sleep32. Furthermore, Unpaired cytokines have been implicated in the modulation of feeding behavior through effects on glial cells33. Glial cells, including those making up the blood-brain barrier (BBB), have recently gained attention for their role in sleep regulation in both flies and mammals34-37. Neurons in the CNS are separated from the circulatory system by the BBB, a selectively semi-permeable cell layer38, which presents a challenge for peripheral hormones to enter and signal to neurons within the brain. However, glial cells within the BBB are ideally positioned to receive and integrate systemic signals from peripheral organs and modulate neuronal function, thereby relaying peripheral information into the brain.

Here, we demonstrate Upd2 and Upd3 cytokine signaling from endocrine EECs in the intestine in Drosophila. Our findings show that Upd2/3 signaling from the EECs to BBB glial cells plays a dual role in sleep regulation. Under normal, healthy conditions, EEC-derived Unpaired signaling sustains wakefulness, whereas in response to oxidative stress that leads to gut inflammation, elevated Unpaired signaling instead promotes sleep. Stress-induced EEC-derived Upd2/3 activates the JAK-STAT pathway in glial cells at the blood-brain interface and adjusts sleep through this activation based on intestinal homeostasis and levels of inflammatory signaling from the gut. Our results suggest that gut-derived Unpaired signaling influences sleep regulation through glial gating of wake-promoting AstA-mediated signals, thus linking intestinal health with CNS-dependent behaviors. These results identify a gut-brain connection by which gut disease impacts sleep regulation.

Results

Gut-derived Unpaired cytokine signaling regulates sleep

To investigate whether cytokine signaling from the gut regulates sleep, we silenced the expression of upd2 and upd3 in the EECs, which are a principal endocrine cell type in the gut that releases signals with systemic effects and constitutes the functional basis of gut-brain signaling. Using voilà-GAL4 (a driver that targets all EECs) to drive RNAi in EECs in conjunction with Tub-GAL80ts (hereafter referred to as voilà>) for temperature-induced RNAi induction exclusively in the adult stage to prevent developmental effects, we observed significant knockdown of the main IL-6 cytokine, upd330, in dissected adult female midguts (Fig. 1a), demonstrating that upd3 is expressed in EECs under normal homeostatic conditions. To eliminate potential neuron-derived phenotypes, we employed R57C10-GAL80, a form of nSyb-GAL80 that effectively inhibits neuronal GAL4 activity, in combination with Tub-GAL80tsand voilà-GAL425,39. To evaluate the effectiveness of this temperature-sensitive EEC-specific driver (referred to as EEC> hereafter) upon adult-restricted induction, we examined midgut upd2 and upd3 transcript levels and observed significant knockdown of both cytokines, which was reproduced with a second independent RNAi line targeting upd3 (Fig. 1b). Importantly, we did not observe any effect of this manipulation on neuronal expression of upd2 or upd3, supporting the specificity of EEC-targeted knockdown (Fig. S1a). Additionally, GFP expressed under the control of an upd3-GAL4 driver containing upd3 enhancer sequences, was apparent in midgut EECs, marked by staining against the EEC fate determinant Prospero (Fig. 1c). To further confirm the expression of upd3 in EECs, we conducted fluorescent in situ hybridization targeting upd3 and prospero in adult midguts. This analysis revealed clear co-localization of upd3 transcripts with pros-positive EECs, consistent with the expression pattern observed using upd3-GAL4–driven GFP in these cells (Fig. S1b). These results show that EECs of the adult female midgut are a source of Upd2 and Upd3 under normal conditions.

EEC-derived Unpaired signaling regulates sleep.

a, upd3 expression levels in midguts expressing RNAi-mediated knockdown of upd3 in EECs using voilà-GAL4 in combination with Tubulin-GAL80ts (voilà>) (N=7). b, upd2 and upd3 expression levels in midguts in animals with EEC knockdown of upd2 or upd3 using voilà-GAL4 in combination with Tubulin-GAL80ts and R57C10-GAL80 (EEC>) (N=4-6). c, Confocal imaging of upd3-GAL4-driven UAS-GFP expression in EECs, co-stained with Prospero (Pros) as an EEC marker (Scale bar, 20 µm). d, Twenty-four-hour sleep profiles in animals with EEC-specific upd2 or upd3 knockdown (N=29-34). e, Total daytime sleep and f, nighttime sleep durations in animals with EEC-specific upd2- or upd3-knockdown flies (N=29-34). g, Twenty-four-hour sleep profiles for global upd2/3 deletion mutants (N=19-32). h, Daytime and i, nighttime sleep durations in global upd2/3 mutants (N=19-32). j, Sleep profiles following EEC-specific CRISPR-mediated upd2 or upd3 knockout (N=24-31). k, Daytime and l, night sleep durations in animals with EEC-specific CRISPR-mediated upd2 or upd3 knockout (N=25-31). m, Twenty-four-hour sleep profiles in animals with AstC-positive-EEC-specific knockdown of upd2 or upd3 using AstC-GAL4 combined with R57C10-GAL80 (AstCGut>) (N=29–30). n, Daytime sleep, and o, nighttime sleep durations in animals with AstCGut>-mediated knockdown of upd2 or upd3 (N=28–32). p, Twenty-four-hour sleep profiles in animals with Tachykinin-positive-EEC-specific knockdown of upd2 or upd3 using Tk-GAL4 combined with R57C10-GAL80 (TkGut>) (N=32). q, Daytime sleep, and r, nighttime sleep durations in animals with TkGut>-mediated knockdown of upd2 or upd3 (N=28–31). Statistical analyses performed using Mann-Whitney tests for panels a and b; ordinary one-way ANOVA with Dunnett’s multiple comparisons for panels e, j, h, i, n, o, q, and r; Kruskal-Wallis ANOVA with Dunn’s multiple comparisons for panels f, k, and l. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

We then explored whether these EEC-derived cytokines govern sleep under such normal conditions, given the known role of cytokines in both healthy and inflammatory states. We found that the knockdown of the two main immune-related cytokines, upd2 or upd3, using voilà> increased the amount of time animals spent asleep (defined as a period of inactivity lasting at least five minutes), especially during the day (Fig. S1c-e). This phenotype was reproducible using the more restricted EEC> driver that includes R57C10-GAL80, suggesting that the EEC-specific loss of these cytokines promotes sleep with pronounced effects during the daytime, when females are typically active (Fig. 1d-f). This outcome was not attributable to off-target effects (two independent RNAi lines targeting the main IL-6-like cytokine, upd3, produced similar phenotypes) or to effects of the RNAi transgenes themselves (Fig. 1d-f and S1f-h). Animals lacking EEC-derived Unpaired signaling also exhibited shorter motion bouts (periods of activity; Fig. S1i,j). However, the EEC-specific knockdown of upd2 or upd3 did not reduce motion-bout activity (the intensity of activity during wake periods), implying that the lack of gut cytokine signaling did not reduce general activity when the animals were awake (Fig. S1k.l), and these effects do not arise from RNAi transgene insertions themselves (Fig. S1m,n). These observations suggest a direct influence of gut-derived Upd2 and Upd3 on sleep rather than a broader impact on general activity levels. Given that sleep and feeding are mutually exclusive behaviors, we measured feeding. We did not detect significant alterations in feeding behavior as a consequence of upd2 or upd3 knockdown in EECs over a 24-hour period using the automated FLIC system (Fig. S1o), nor did we observe an effect of EEC-specific upd3 overexpression on food consumption via a dye assay (Fig. S1p). Furthermore, EEC-specific upd3 knockdown did not affect the animals’ metabolic state as reflected in their levels of stored triacylglyceride (TAG) (Fig. S1q). Therefore, the sleep phenotype exhibited by animals with EEC-specific upd2/3 knockdown is not associated with changes in metabolism, appetite, or feeding behavior.

To further explore the role of Upd2 and Upd3 in sleep regulation, we made use of deletion mutations that disrupt upd3 alone (upd3τ1) or in combination with upd2 (upd2,3τ1). Both mutant lines exhibited a pronounced increase in sleep (Fig. 1g-i), with strong effects on daytime sleep, phenocopying the RNAi-mediated knockdown in the EECs. We additionally disrupted the upd2 or upd3 genes specifically in the EECs using somatic tissue-specific CRISPR-mediated deletion. EEC-specific CRISPR-mediated knockout of upd2 or upd3, induced by UAS-controlled gRNA pairs designed to excise portions of each gene’s coding sequence, led to significantly elevated sleep (Fig. 1j-l), further reinforcing these cytokines’ role in sleep modulation.

To further dissect the cellular source of gut-derived cytokines regulating sleep, we analyzed the contribution of the two major EEC populations in the adult Drosophila midgut, marked by expression of either Allatostatin C (AstC) or Tachykinin (Tk)20. These two molecularly defined groups encompass the vast majority of EECs. We used AstC-GAL425 and Tk-GAL424 drivers, which are knock-in lines carrying GAL4 inserted at the endogenous AstC or Tk loci, thereby enabling precise genetic targeting of EECs based on their native hormone expression profile. To restrict GAL4 activity to the gut and thus avoid effects from neuronal expression, both drivers were combined with R57C10-GAL80, generating AstCGut-GAL4 (AstCGut>) and TkGut-GAL4 (TkGut>) drivers. Using these tools, we selectively knocked down upd2 or upd3 in either the AstC-positive EECs or the Tk-positive cells. Knockdown of either cytokine in AstC-positive EECs significantly increased sleep (Fig. 1m-o), phenocopying the effect observed with knockdown in all EECs (Fig. 1d-f). In contrast, knockdown of upd2 or upd3 in Tk-positive EECs had no significant effect on sleep (Fig. 1p-r). These findings indicate that AstC-positive EECs are a major source of sleep-regulating Unpaired cytokines, whereas Tk-positive EECs do not appear to contribute significantly to this function. Consistent with this, we also observed effective knockdown of upd3 transcripts in dissected midguts using the AstCGut> driver, indicating that upd3 is endogenously expressed in the AstC-positive EEC population (Fig. S1r). Collectively, these findings demonstrate that Upd2 and Upd3 expressed by EECs are important modulators of diurnal sleep patterns in Drosophila under normal homeostatic conditions, and they further identify AstC-positive EECs as a key cellular source of these sleep-regulating cytokines.

Glial cytokine JAK-STAT signaling regulates sleep

To identify the CNS targets of EEC-derived Upd2/3 cytokine signaling by which they regulate sleep, we examined the effects of targeted knockdown of the Upd2/3 JAK-STAT-linked receptor dome in neurons or glia, the two main cell types in the CNS. Pan-neuronal dome knockdown using the driver R57C10-GAL425 did not significantly alter daytime or nighttime sleep in adult females (Fig. 2a), thus failing to recapitulate the sleep increase observed upon loss of upd2 or upd3 in EECs (Fig. 1). This suggests that neurons are not the targets by which gut Unpaired cytokine signaling regulates sleep. In contrast, knockdown of dome in glial cells using the pan-glial driver repo-GAL4 (repo>) resulted in a pronounced increase in daytime and nighttime sleep (Fig. 2a), similar to the phenotype observed upon EEC-specific loss of Unpaired cytokines. To substantiate this observation further, we silenced dome expression in all glia using three independent RNAi lines, all of which strongly induced sleep during the day, effectively ruling out any off-target or transgene-background effects (Fig. 2b-d and S2a-e). Additionally, animals with glial dome knockdown displayed shorter daytime motion bouts, suggesting reduced periods of wakefulness (Fig. 2e and S2f), without exhibiting decreased activity during these bouts (Fig. 2f and S2g), phenocopying the effects of upd2/3 knockdown in EECs (Fig. S1i,k). We also assessed whether glial-specific dome knockdown might affect feeding and energy storage, but we observed no reduction in food intake (Fig. S2h), and no changes in TAG levels were detected (Fig. S2i). These findings collectively argue that augmented sleep resulting from impaired JAK-STAT signaling in glia is not due to a general decline in activity but rather represents a specific regulation of sleep itself.

EEC-derived Unpaired signaling regulates glial JAK-STAT activity that modulates sleep.

a, Day and night sleep measurements for flies with knockdown of IL-6 related Unpaired cytokine receptor domeless (dome), which activates JAK-STAT, in neurons (R57C10-GAL4, R57C10>) and glial cells (repo-GAL4, repo>) (N=25-39). b, Twenty-four-hour sleep profiles for controls and animals with glia-specific dome knockdown (N=25-32). c, Total day sleep duration and d, total night sleep duration for animals with glia-specific dome knockdown and control flies (N=25-32). e, Motion-bout length and f, motion-bout activity in animals with glia-specific dome knockdown and controls (N=25-32). g, Representative images of brains of from controls and animals with EEC knockdown upd2 or upd3 using voilà>, with 10xSTAT-GFP. GFP expression (green) reflects JAK/STAT activity, and Repo labeling (red) indicates glial cells (Scale bar, 50 µm). Insets show zoomed views of STAT-GFP+ and Repo+ glial cells (Scale bar, 15 µm). h, Quantitative analysis of GFP intensity in the layer of glial cells located at the surface of the brain in animals with EEC knockdown of upd2 or upd3 and controls (N=8). Statistical analyses were conducted using parametric t-tests for panel a; Kruskal-Wallis ANOVA with Dunn’s multiple comparisons for panels c, d, e, and f; and ordinary one-way ANOVA with Dunnett’s multiple comparisons for panel h. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

To directly assess the functional Unpaired-mediated communication between the gut and glial cells, we manipulated upd2 and upd3 in EECs of animals carrying a ubiquitous transgenic JAK-STAT reporter (10xSTAT-GFP)40. Knockdown of either upd2 or upd3 in the EECs led to a marked decrease in JAK-STAT reporter activity within Repo-positive glial cells under normal conditions (Fig. 2g,h), suggesting that Upd2/3-mediated signaling from the EECs to the brain’s glial cells activates JAK-STAT signaling. Taken together this suggests that gut-to-glia communication via Upd2 and Upd3 modulates diurnal sleep patterns through glial JAK-STAT activation and that these cytokines are required for the maintenance of wakefulness during the day under healthy conditions.

Oxidative stress modulates sleep through gut-derived cytokine signaling

Having established the significance of gut-derived Unpaired cytokines in maintaining wakefulness under normal conditions, we next explored their role in sleep regulation in the context of gut disturbances that trigger immune and inflammatory responses. Gut infection leads to increased ROS levels and induces local cytokine production26,27,41, and oxidative stress is a key feature of inflammatory conditions in the intestine42. Dietary H2O2 treatment leads to local intestinal responses comparable to those observed during pathogenic challenges43, suggesting that H2O2 feeding provides a controlled method to elevate intestinal ROS levels and examine the specific effects of ROS-induced cytokine signaling. We used this paradigm to ask whether intestinal oxidative stress might elevate the levels of Upd2 and Upd3 in the gut, and we found substantial upregulation of upd3 expression in dissected midguts of females challenged with ROS by feeding 1% H2O2 adult-specific, cornmeal-free diet for 20 hours (Fig. 3a). This effect mirrored the upregulation observed with EEC-specific overexpression of upd3, indicating that it reflects physiologically relevant production of Upd3 by the gut in response to oxidative stress (Fig. 3a). Oxidative stress also promoted upd2 expression, albeit to a lesser extent, and this effect was not modulated by simultaneous EEC-specific upd3 overexpression.

EEC-derived Unpaired signaling modulates sleep in response to intestinal oxidative stress.

a, Measurement of upd2 and upd3 expression in the midgut upon 20-hour treatment with 1% H2O2-laced food or with overexpression of upd3 (upd3-OE) using EEC> (N=4). b and c, Assessment of sleep duration over consecutive days during the daytime (ZT0-ZT12) and nighttime (ZT12-ZT24) in animals exposed to food containing (b) 0.1% H2O2 (N=26-31) or (c) 1% H2O2 (N=23-29). d, Daily sleep, and e, nightly sleep amounts measured over one day under standard food conditions followed by two consecutive days on 1% H2O2-containing food in animals with EEC-specific upd2 or upd3 knockdown and controls (N=23-30). Experiments measuring sleep levels in controls and animals lacking EEC-derived upd2 or upd3 were performed concurrently and share the “control” data, but results are presented in separate figures (b, c, d, and e) for clarity. In d, two-way ANOVA revealed significant genotype x diet interactions for upd2-i (p = 0.0076), upd3-iKK (p = 0.0003), upd3-iTRiP (p = 0.0204), and upd3-iGD (p = 0.0040), relative to the control, indicating that the sleep response to oxidative stress depends on EEC-derived Unpaired signaling. f, Sleep profiles and measurements of daytime (g) and nighttime (h) sleep across a two-day period, encompassing one day on standard diet followed by one day on 1% H2O2-laced food to induce oxidative stress, in flies with AstC-positive-EEC-specific knockdown of upd2 or upd3 using AstCGut> compared to controls (N=31–32). i, Survival rates under a 1% H2O2-induced oxidative stress diet in controls and animals with EEC-specific upd2 or upd3 knockdown (N=23-30). j, A 48-hour sleep profile comparison between global upd2/3 mutants and w1118 controls under one day of standard food conditions followed by one day of 1% H2O2-induced stress (N=18-63). k and l, Quantification of daytime and nighttime sleep durations in upd2/3 mutants versus w1118 controls under normal-food conditions and the following day exposed to food containing 1% H2O2 (N=18-63). In k, two-way ANOVA showed significant genotype × diet interaction (p < 0.0001), confirming a role for Unpaired cytokines in ROS-induced sleep modulation. m, Observation of sleep patterns, and n, measurements of daytime sleep across a three-day period, encompassing a day on standard diet, subsequent day on 1% H2O2-laced food to induce oxidative stress, and a final day back on standard diet to monitor recovery, in flies with EEC-specific overexpression of upd3 (upd3-OE) compared to controls (N=29-32). Statistical analyses were performed using Kruskal-Wallis ANOVA with Dunn’s multiple comparisons for panels a, b, c, d, e, i, and n; Mann-Whitney test for panels g, h, k, and l. Interaction effects were assessed using two-way ANOVA where indicated. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

We next investigated whether sleep is modulated by intestinal oxidative stress and if Unpaired signaling from EEC is required for this response. We induced intestinal oxidative stress by exposing animals to a diet supplemented with H2O2 at Zeitgeber Time 0 (ZT0), the onset of the light phase, in a 12-hour light/dark cycle. Exposure to a lower H2O2 concentration (0.1%) incrementally increased daytime sleep amount over successive days (Fig. 3b). In contrast, a higher H2O2 concentration (1%) triggered an immediate augmentation of daytime sleep (Fig. 3c). Additionally, to ensure that the observed sleep increase was due to the presence of H2O2 itself rather than the procedure of food supplementation, we conducted a control experiment in which animals were fed standard food prepared using the same protocol and replaced daily, but without H2O2. These animals did not exhibit increased sleep, confirming that the effect is attributable to intestinal ROS (Fig. S3a).

These observations suggest that intestinal oxidative stress dose-dependently modulates sleep. Since 1% H2O2 feeding induced robust responses both in upd3 expression and in sleep behavior, we asked whether gut-derived Unpaired signaling might be essential for the observed ROS-induced sleep modulation. Indeed, EEC-specific RNAi targeting upd2 or upd3 abolished the sleep response to 1% H2O2 feeding. Animals with EEC-specific knockdown of upd2 or upd3 did not exhibit increased daytime sleep in response to the induction of oxidative stress in the intestine, even over two consecutive days of exposure to 1% H2O2-containing diet (Fig. 3d,e). The specificity of this response was corroborated by three independent RNAi lines targeting upd3, negating the possibility of RNAi off-target effects (Fig. 3d,e), and the loss of response to ROS was also not attributable to the transgenes themselves (Fig. S3b-e). Intriguingly, animals lacking upd3 in the EECs not only did not increase their sleep under oxidative stress but indeed appeared to lose nighttime sleep in response to enteric stress. Moreover, knockdown of upd2 or upd3 limited to the AstC-positive EEC subpopulation still prevented the H2O2-induced increase in sleep (Fig. 3f-h). These findings indicate that Unpaired signaling from AstC-positive EECs is necessary for mediating the sleep response to intestinal oxidative stress and highlight a specific EEC subtype as a critical source of cytokine signaling in this context.

We next tested whether this sleep phenotype might be associated with general physiological processes rendering animals lacking EEC Unpaired signaling more susceptible to ROS-induced damage. However, when we assessed survival on 1% H2O2-containing food, animals with upd2 or upd3 knockdown in EECs displayed no additional sensitivity to oxidative stress, compared to controls (Fig. 3i). To further rule out nonspecific toxicity, we examined whether 1% H2O2 feeding under our experimental conditions causes neuronal damage. Using a TUNEL assay for apoptosis, we found no evidence of increased neuronal cell death in animals fed 1% H2O2 for 24 hours, suggesting that the observed sleep phenotypes are not attributable to general neuronal toxicity (Fig. S3f,g). This indicates that the loss of EEC-derived Unpaired signaling specifically leads to an impaired behavioral sleep response to intestinal oxidative stress, rather than to compromised physiological processes that would make the animals more vulnerable to oxidative-stress insults. We also examined whether animals lacking EEC-derived Unpaired signaling exhibit normal behavioral responses to other conditions that modulate sleep, which would suggest a specific requirement for this signaling in responding to intestinal oxidative stress. Animals typically suppress their sleep in response to nutritional deprivation, a behavior conserved across species that is believed to facilitate food-seeking activities and that is also influenced by EEC-mediated hormone signaling25,44. Animals lacking EEC-derived upd3 suppressed their sleep to a similar extent as controls in response to starvation, indicating a normal sleep response to nutritional stress (Fig. S3h-j).

Although we observed behavioral phenotypes with manipulations of either upd2 or upd3 alone, suggesting that both are required for normal function, Upd2 and Upd3 likely function at least partially redundantly or additively in their regulation of sleep, as is the case for other processes45. Moreover, RNAi effects do not result in a complete loss of function. Therefore, we speculated that a stronger disruption and combined knockout of both upd2 and upd3 might lead to even more pronounced phenotypes. We therefore tested the upd3Δ single-deletion line and the upd2,3Δ double-deletion mutants. Whereas upd3Δ and upd2,3Δ mutants exhibited increased baseline sleep under homeostatic conditions, these animals not only failed to increase their sleep in response to oxidative stress but indeed showed a strong reduction in daytime and nighttime sleep under oxidative-stress conditions (Fig. 3j-l). These results suggest that contrary to its role in promoting wakefulness during normal homeostatic conditions, the enhanced ROS-induced Unpaired signaling from EECs helps sustain a higher sleep level during periods of oxidative stress. This indicates a dual functionality of Unpaired cytokine signaling, in which low Unpaired signaling promotes wakefulness under normal conditions, whereas higher ROS-induced Unpaired signaling facilitates a shift to restorative sleep during intestinal stress.

We therefore investigated whether higher levels of Unpaired signaling from the gut, comparable to the level produced during oxidative stress, could enhance sleep in the absence of exogenous stressors. We analyzed the effect of upd3 overexpression in EECs, which drives expression of midgut upd3 to levels similar to those induced by 1% H2O2 feeding (Fig. 3a). Consistent with a model in which high levels of Upd3, like those that would occur during periods of elevated intestinal oxidative stress, promotes daytime sleep, animals overexpressing upd3 in the EECs exhibited increased sleep during the day, even in the absence of H2O2-induced oxidative stress (Fig. 3m,n). These animals further increased their sleep in response to H2O2-induced enteric oxidative stress, unlike those lacking gut-derived upd3 (Fig. 3d,e), suggesting they remain able to mount an additional ROS-induced Unpaired signaling response on top of the overexpression-induced levels. After the animals were switched back to normal food after one day of oxidative stress, both control animals and those with EEC-specific upd3 overexpression exhibited even more sleep than during the previous day under oxidative-stress conditions (Fig. 3m,n). This suggests a robust recovery sleep response following the insult, likely mediated by Unpaired signaling, since the effect is more pronounced with upd3 overexpression. Taken together our results show that control animals increase their sleep during oxidative stress, likely as an adaptive recovery response. In contrast, animals with EEC-specific knockdown of unpaired cytokines do not exhibit this ROS-induced sleep response; instead, they experience sleep loss under such conditions. This suggests that while Unpaired signaling promotes wakefulness during normal healthy conditions, temporary ROS-induced elevation of gut Unpaired signaling suppresses arousal and leads to more sleep.

EEC-derived Unpaired cytokine signaling activates glial JAK-STAT under oxidative stress

To investigate whether oxidative stress enhances glial JAK-STAT signaling and, if so, whether this enhancement might be mediated by gut-derived Upd2 and Upd3, we assessed glial JAK-STAT reporter activity using the dual-color TransTimer system, which provides temporal information about JAK-STAT signaling46. In this system, active STAT promotes the expression of a construct encoding a short-lived destabilized GFP (dGFP, half-life ∼2 hours) and a long-lived RFP (half-life ∼20 hours) separated by a 2A peptide (6xSTAT-dGFP:2A::RFP); a higher ratio of GFP to RFP in a given cell reflects more recent JAK-STAT signaling. We explored whether JAK-STAT signaling responds dynamically to intestinal oxidative stress and assessed two daily time points. In control animals, we observed no circadian changes between ZT0 (lights on) and ZT12 (lights off) in the superficial layer of cells surrounding the brain (Fig. 4a,b), which is composed of glia47. However, we observed a significant increase in GFP signal at ZT0 in animals fed for 20 hours with 1% H2O2-containing food, indicating recent JAK-STAT activity in the surface glia. Next, we investigated whether gut-derived Unpaired signaling is responsible for this upregulation by combining the 10xSTAT-GFP reporter with knockdown of upd2 or upd3 in the EECs. Whereas glial JAK-STAT reporter activity was upregulated by oxidative stress (20 hours’ 1%-H2O2 feeding) in control animals, this response was abolished in animals with EEC-specific knockdown of upd2 or upd3, indicating that this response is dependent on these EEC-derived cytokines (Fig. 4c,d). Since in this case we used voilà> without the pan-neuronal R57C10-GAL80 element to limit knockdown in EECs, we measured the expression of upd2 and upd3 in heads to check for any unintended neuronal effects that might contribute to the observed effect on glial JAK-STAT activity. We detected no changes in the expression of these genes in the head, confirming that the observed JAK-STAT activation in glial cells is attributable to cytokines derived from EECs (Fig. S4a,b). To test the ability of gut-derived Upd3 to drive events in the brain in another way, we made use of cells’ homeostatic response to changes in signaling input. Receptor expression is often upregulated in response to low levels of a ligand as a compensatory mechanism to enhance cellular sensitivity48. We observed an upregulation of dome transcript levels in the heads of animals with EEC-specific knockdown of upd3, exposed to oxidative stress induced by 15 hours of feeding with food laced with 1% H2O2 (Fig. 4e). Increased dome expression suggests reduced Unpaired ligand availability as a result of the loss of EEC-derived Upd3. Together, our results demonstrate that EEC-derived Unpaired cytokine signaling is required for activating glial JAK-STAT under oxidative stress.

Activation of glial JAK-STAT signaling by EEC-derived Unpaired cytokines in response to enteric oxidative stress.

a, Representative images of brains from flies expressing the STAT-::dGFP::2A::RFP reporter, where green (dGFP) reflects recent JAK-STAT activity due to its rapid degradation, and purple (RFP) indicates longer-term pathway activation due to its higher stability. Left panels show brains at lights-on (ZT0), middle panels show brains at lights off (ZT12), and right panels depict brains after 20 hours of oxidative stress induced by 1% H2O2-containing food, imaged at lights-on time ZT0. White dotted lines outline the brain perimeter. Scale bar, 50 µm. b, Ratio of dGFP to RFP fluorescence intensity at ZT0, at ZT12, and after oxidative stress (at ZT0), as depicted in panel a, to show dynamic changes in JAK-STAT activity (N=108-461, indicating the number of cells counted). c, Representative images of brains displaying 10xSTAT-GFP expression under homeostatic conditions and after oxidative stress in control flies (“Ctrl”) and flies with EEC-specific upd2 or upd3 knockdown. Scale bar, 50 µm. Inset panels provide magnified views of glia cells labeled by anti-Repo. Scale bar, 15 µm. d, Quantification of 10xSTAT-driven GFP intensity in glial cells under homeostatic and oxidative-stress conditions, demonstrating the impact of EEC-specific cytokine knockdown (N=8, indicating the number of brains). e, qPCR analysis of dome expression in the brains of flies with EEC-specific upd3 knockdown in comparison to voilà> controls (N=5). Statistical analyses were conducted using Kruskal-Wallis ANOVA with Dunn’s multiple comparisons for panel b; ordinary one-way ANOVA with Tukey’s multiple comparisons for panel d; and two-sided unpaired t-tests for panel e. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

Glia JAK-STAT modulates sleep in response to oxidative stress

Since EEC-derived Unpaired signaling promotes oxidative stress-induced sleep and glial JAK/STAT activity, we investigated whether the observed glial JAK-STAT signaling is involved in the modulation of sleep in response to intestinal oxidative stress. Knockdown of dome in all glial cells using the repo> driver completely abolished the ROS-induced daytime-sleep response when animals were the fed 1%-H2O2 supplemented diet. The specificity of this effect was confirmed using three RNAi lines and with transgenic RNAi controls (Fig. 5a and S5a-c). Mirroring the effects observed with upd2,3Δ mutants (Fig. 3g-i), glia-specific dome knockdown (p < 0.00001 for dome-iKK and p = 0.0556 for dome-iTRiP) resulted in progressive and substantial sleep loss over two consecutive days on 1% H2O2-containing food. To rule out developmental effects, we restricted glial knockdown of dome to the adult stage using the repo> driver in combination with Tub-GAL80ts (repoTS>) and observed similar effects (Fig. 5b).

EEC-derived Unpaired and glial Domeless signaling modulate sleep during intestinal oxidative stress.

a, Daytime sleep duration in flies with glia-specific dome knockdown under standard and oxidative-stress conditions induced by 1% H2O2 in food (N=25-32). Two-way ANOVA revealed significant genotype × diet interaction (p < 0.0001), indicating that glial Domeless is required for sleep regulation during oxidative stress. b, Daytime sleep duration in flies with repo-driven dome knockdown restricted to the adult stage using Tub-GAL80ts (repoTS>) under normal conditions and during exposure to 1% H2O2-containing food (N=29-32). Two-way ANOVA showed a significant genotype × diet interaction (p < 0.0001), further supporting a role for glial dome in regulating sleep in response to gut oxidative stress. c, Daytime sleep during a three-day period, encompassing a day on standard diet, subsequent day on 1% H2O2-laced food to induce oxidative stress, and a final day back on standard diet to monitor recovery, in controls and animals with glia-specific dome knockdown (N=24-32). Two-way ANOVA revealed significant genotype × diet interaction (p < 0.0001). d, Survival rates of controls and flies with glial-specific dome knockdown after exposure to oxidative stress by 1% H2O2-laced food (N=31). e, Sleep profiles, and f, daytime sleep duration for animals with EEC-specific upd3 knockdown compared to control flies across a 36-hour period encompassing 24 hours on standard diet followed by 12 hours on oxidative-stress conditions induced by 4% H2O2-containing food (N=15-30). Two-way ANOVA showed a significant genotype × diet interaction (p < 0.0001). g, Sleep profiles, and h, daytime sleep duration for animals with glia-specific dome knockdown compared to control flies across a 36-hour period encompassing 24 hours on standard diet followed by 12 hours under oxidative-stress conditions induced by 4% H2O2-containing food (N=20-32). Two-way ANOVA revealed significant genotype × diet interaction (p < 0.0001). i, Nighttime sleep durations for animals under 4% H2O2 oxidative-stress conditions in controls and animals expressing adult-restricted knockdown of dome in glia (N=31-32). Two-way ANOVA revealed significant genotype × diet interaction (p < 0.0001). Statistical tests: Kruskal-Wallis ANOVA with Dunn’s multiple comparisons for panels a, c, and d; Unpaired two-sided t-tests for panels b, f, and h; and Mann-Whitney test for panel i. Interaction effects were assessed using two-way ANOVA where indicated. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

We then examined the dynamics of sleep-regulatory glial JAK-STAT signaling by inducing oxidative stress for one day and then transferring the animals to normal food to observe the recovery response. The results showed that, in response to 1% H2O2-containing food, animals lacking glial dome expression displayed a sleep response opposite from that of controls, with a reduction in sleep duration rather than an increase, confirmed using independent RNAi lines (Fig. 5c). This phenotype is similar to that seen in the upd2,3Δ double mutants (Fig. 3h-j). During the recovery phase, after the animals had been switched back to normal food, the sleep level of controls increased even further, but the sleep duration exhibited by animals with glia-specific dome knockdown immediately reverted to pre-stress levels. This pattern indicates that animals with inhibited glial JAK-STAT signaling display an aberrant dynamic sleep response to oxidative stress that is not a consequence of a physiological breakdown but rather arises from altered inhibitory sleep-regulating mechanisms. In line with this and paralleling the loss of upd2 or upd3 in the EECs, dome knockdown in glial cells did not decrease survival on H2O2-containing food (Fig. 5d). This supports the notion that physiological resistance to oxidative stress remains unaltered by gut-glial Unpaired signaling, which in turn indicates that the signaling modulation leads to a specific sleep phenotype. This is likely an important adaptive response under natural conditions, promoting recovery and maintaining homeostasis during or after transient stress episodes.

We further assessed whether glial loss of dome affected homeostatic sleep responses induced by sleep deprivation by evaluating the animals’ ability to recover sleep after deprivation occurring during the second half of the night (ZT18-ZT24). Like controls, animals with glial-specific dome knockdown exhibited increased sleep (rebound sleep) in the morning hours (ZT0-ZT2) following sleep deprivation (Fig. S5d,e). This indicates that they exhibit normal rebound sleep responses to deprivation and retain the capability to further increase their sleep. Collectively, these data suggest that Dome-mediated JAK-STAT signaling in the glial cells specifically regulates ROS-induced sleep responses.

We next investigated whether increased intestinal oxidative stress would exacerbate the phenotypes associated with the loss of upd3 in EECs or dome in glial cells by exposing the animals to food containing 4% H2O2 and observing changes in their sleep architecture. Oxidative stress resulted in increased sleep in control animals, as anticipated (Fig. 5e-h). However, in animals with EEC-specific upd3 knockdown, or glia-specific dome RNAi, exposure to 4% H2O2-containing food led to a pronounced loss of sleep during the daytime. For the EEC-specific upd3 knockdown, the RNAi effect was induced at the adult stage (Fig. 5e,f). We therefore also confirmed that adult-restricted knockdown of dome in glial cells resulted in similar phenotypes (Fig. 5i). Thus, under conditions of intensified intestinal stress induced by 4% H2O2 in the food, the loss of upd3 in EECs phenocopies the glial knockdown of dome, leading to reduced sleep and increased wakefulness.

BBB glial JAK-STAT pathway activation drives sleep in response to intestinal oxidative stress

To determine the subset of glial cells responsible for mediating Unpaired-driven sleep regulation, we focused on the perineurial and subperineurial glial cells that form the BBB. These BBB glial cells serve as the interface between the CNS and the periphery, including its organs47, and are ideally situated to receive circulating signals from the intestine. Using the 10xSTAT-GFP reporter, we assessed whether Upd3 from the gut activates JAK-STAT signaling within BBB glial cells. Knockdown of upd3 in the EECs using voilà> (without R57C10-GAL80), which in this assay drives specific knockdown in the gut without detectable neuronal effects (Fig. S4a,b), resulted in decreased GFP intensity in the outermost glial cell layer of the central brain facing the periphery after 15 hours of exposure to 1% H2O2-containing food, indicating reduced JAK-STAT activity in these cells (Fig. 6a,b). Interestingly, this knockdown did not affect JAK-STAT activity in the outer glial layer of the ventral nerve cord (VNC), suggesting that Upd3 acts specifically on brain BBB glia in response to intestinal oxidative stress.

BBB glia drive Domeless-mediated sleep responses to intestinal oxidative stress.

a, Representative images showing GFP expression driven by 10xSTAT-GFP in controls and animals with voilà-GAL4 (voilà>)-driven upd3 knockdown in EECs under intestinal ROS induced by 20 hours’ exposure to 1%-H2O2-containing food. The top panels depict overall brain and ventral nerve cord (VNC) structure with views of surface or deeper layers; the bottom panels provide zoomed-in views, highlighting the BBB glia at the interface between the brain and external environment. Dotted lines indicate brain and VNC perimeters. Scale bars, 50 µm (top) and 15 µm (bottom). b, Quantification of GFP intensity in the brain and VNC in BBB glia in controls and animals with EEC knockdown of upd3 under ROS stress, induced by exposure to 1% H2O2-laced food (N=7). c, Daytime, and d, nighttime sleep durations in flies with BBB-glia-specific knockdown of dome or overexpression of hopTumunder normal conditions, during exposure to 1% H2O2-containing food, and subsequent recovery on normal diet (N=23-32). In c, two-way ANOVA revealed significant genotype × diet interaction (p < 0.0001), indicating that BBB-glial Domeless is required for daytime sleep regulation under oxidative stress. e, Daytime, and f, nighttime sleep durations in flies with BBB-glia-specific knockdown of dome or overexpression of hopTum under normal conditions, during exposure to 4% H2O2-containing food, and subsequent recovery on normal diet (N=23-32). In e, two-way ANOVA revealed significant genotype × diet interaction (p < 0.0001), confirming the importance of BBB-glial Domeless signaling during higher levels of oxidative stress. Statistical tests used: Unpaired two-sided t-tests for panel b; Kruskal-Wallis ANOVA with Dunn’s multiple comparisons in panels c, d and f; ordinary one-way ANOVA with Dunnett’s multiple comparisons for panel e. Interaction effects were assessed using two-way ANOVA where indicated. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

We further examined BBB-specific JAK-STAT signaling-mediated effects on sleep by specifically manipulating the subperineurial glial cells – those that form the permeability barrier – using moody-GAL4 (moody>). As observed with other manipulations, knockdown of dome in these BBB glial cells led to increased sleep during normal homeostatic conditions (Fig. 6e,f). Control animals exhibited the expected sleep increase both during 1%-H2O2 exposure and during the recovery period, but loss of dome in the subperineurial BBB glia blocked these effects. When oxidative stress levels were elevated further using 4% H2O2-containing food, dome-RNAi in BBB glial cells led to sleep loss in response to oxidative stress, with sleep levels rebounding to pre-stress levels on the subsequent recovery day when animals were returned to a normal diet (Fig. 6e,f). These results indicate that disrupting dome specifically in the subperineurial glial cells of the BBB recapitulates the phenotypes observed with pan-glial dome knockdown or with EEC-specific upd3 knockdown, and they suggest that Dome-mediated JAK-STAT activation in the subperineurial BBB cells is required for maintaining an increased sleep state during intestinal oxidative stress.

To assess the sufficiency of subperineurial JAK-STAT signaling in inducing sleep, we activated the pathway in these cells by expressing a hyperactivated variant of the Drosophila JAK ortholog Hopscotch (HopTum). Expressing this protein in BBB glia led to increased sleep under normal conditions (without unusual oxidative stress), consistent with a sleep-promoting effect of high JAK-STAT signaling (Fig. 6c-f). Moreover, animals with overactive JAK signaling in BBB glia exhibited a further increase in sleep both during oxidative stress and in the subsequent recovery phase, in contrast to the effects seen with dome knockdown. This suggests that the combined activation of JAK-STAT induced by intestinal ROS and expression of HopTum leads to additive increases in sleep. Collectively, our data indicate that JAK-STAT signaling specifically in the subperineurial glial of the BBB links sleep responses to intestinal oxidative stress.

AstA signaling promotes wakefulness and mediates ROS-induced sleep regulation in BBB glia

Our results indicate that the effect of gut-to-glia Unpaired cytokine signaling is both dose-dependent and context-dependent. During intestinal oxidative stress, ROS-induced EEC Unpaired signaling leads to high JAK-STAT activity in subperineurial glial cells. Given that animals lacking this gut cytokine-to-glial signaling fail to maintain a high sleep state during oxidative stress, instead exhibiting increased wakefulness, this pathway appears to suppress wake-promoting signals under such conditions. The role of such wake-suppressive effects is likely to enhance sleep, aiding the process of recovery from intestinal damage. To identify potential wake-promoting signals that might be gated by JAK-STAT signaling, we examined a published dataset of genes whose expression in glia is positively or negatively correlated with these cells’ JAK-STAT activity following enteric infection33. The receptors for Allatostatin A (AstA), AstA-R1 and AstA-R2, both ranked among the top 4% of genes most strongly downregulated by JAK/STAT signaling (with AstA-R1 expression reduced by ∼80% and AstA-R2 by ∼90%). Notably, these were the only peptide-hormone G-protein-coupled receptors (GPCRs) downregulated in the JAK-STAT-activate glial cells. This suggests that upon intestinal infection, the JAK-STAT pathway is activated in glial cells, which suppresses AstA signaling by reducing the expression of the AstA receptors. Considering the central role of neuronal AstA in sleep-regulatory circuits49,50, we investigated whether AstA might constitute a wake-promoting signal that is inhibited in glial cells by gut-derived Unpaired signaling. To evaluate the expression pattern of AstA-R1 and AstA-R2 within glial populations, we employed AstA-R1-GAL4 and AstA-R2-GAL4 knock-in constructs to drive the expression of nuclear-localized RFP. We co-stained the brain with antibodies against the glial transcription factor Repo, which marks the nuclei of glial cells. We observed that the outer layer of glial cells at the barrier between the brain and the periphery – constituting the BBB – express both AstA-R1 and AstA-R2 (Fig. 7a). These findings are in line with previously reported data showing expression of AstA-R1 and AstA-R2 in glial cells referred to above33. To functionally characterize the role of AstA signaling in these cells, we knocked down AstA-R2 in BBB glial using moody>. This led to a significant reduction in AstA-R2 transcript levels in dissected brains, indicating that BBB glia is a significant source of AstA-R2 expression (Fig. S6a). In support of an inhibitory role of Unpaired signaling, we observed that AstA-R1 and AstA-R2 expression was upregulated in the heads of animals with EEC-specific upd3 knockdown, following gut-oxidative stress induced by feeding with 1% H2O2-laced food for 20 hours (Fig. 7b). To demonstrate that this is caused by a failure to suppress AstA receptors in glial cells, we examined AstA receptor expression in brains following glia-specific dome knockdown in animals fed 1% H2O2-containing food for 20 hours. Indeed, glial-specific dome knockdown led to strong upregulation of both AstA-R1 and AstA-R2, indicating that ROS-induced glia-mediated Unpaired signaling is inhibiting AstA receptor expression (Fig. 7c).

Gut Unpaired cytokine signaling inhibits wake-promoting AstA signaling.

a, Images of brains from animals with AstA-R1-GAL4 and AstA-R2-GAL4 driving nuclear dsRed (nRFP, magenta) and co-stained with anti-Repo antibodies (green) show glial cells. The yellow dashed line indicates the interface between the brain and the external space, with the area below housing the (BBB glial cells (Scale bar, 10 µm). The yellow demarcation accentuates the separation between the cerebral interior and the external milieu, identifying the location of BBB glial cells underneath this partition (Scale bar is 10 µm). b, Relative expression of AstA-R1 and AstA-R2 in heads from animals with upd3 knockdown in EECs driven by voilà-GAL4 (voilà>) compared to the control group after 20 hours on 1% H2O2-laced food to induce oxidative stress (N=5-6). c, Relative expression of AstA-R1 and AstA-R2 in brains from animals with dome knockdown in glial cells driven by repo-GAL4 (repo>) compared to the control group after 20 hours on 1% H2O2-laced food to induce oxidative stress (N=6). d, Sleep patterns, and e, daytime sleep across a three-day period, encompassing a day on a standard diet, a subsequent day on 1% H2O2-laced food to induce oxidative stress, and a final day back on the standard diet to monitor recovery, in flies with BBB-glia-specific knockdown of AstA-R2 compared to control (N=28-32). In e, two-way ANOVA revealed significant genotype × diet interaction (p = 0.0114, supporting a role for glial AstA-R2 in ROS-induced sleep regulation. f, AstA transcript levels in brains and midguts from controls and animals with knockdown of AstA in AstA+ EECs using AstA-GAL4 (AstA>) in combination with R57C10-GAL80 to suppress neuronal GAL4 activity, referred to as AstAGut> (N=5-6). g, Sleep profiles, and h, daytime sleep on standard food of animals with AstA knockdown in AstA+ EECs using AstAGut> and controls (N=30-32). i, Sleep profiles, and j, daytime sleep on standard food of controls, animals with TrpA1-mediated activation of AstA+ EECs, and animals with TrpA1-mediated activation of AstA+ EECs with simultaneous knockdown of AstA (N=30-32). k-m, Quantification of AstA transcript levels in whole midguts (k) and AstA peptide levels in the R5 region of the posterior midgut (l) on standard diet, after 1 day on 1% H2O2-laced food to induce oxidative stress, and during recovery following H2O2 exposure (k: N = 5–6; l: N = 126–170). m, Representative images of R5 regions stained with anti-AstA antibody (scale bar: 25 µm). n, Diagram illustrating the role of enteroendocrine cells (EECs) in regulating wakefulness and sleep through Unpaired cytokine signaling under homeostatic and stress conditions. Left: Under homeostatic conditions, EECs release baseline levels of Unpaired, which interacts with the blood-brain barrier (BBB) to maintain normal JAK-STAT signaling and AstA transduction, promoting wakefulness. Right: In response to stress and disease, reactive oxygen species (ROS) increase in EECs, leading to elevated release of Unpaired. This surge in Unpaired upregulates JAK-STAT signaling in BBB glia, which inhibits wake-promoting AstA signaling by suppressing AstA receptor expression, thus resulting in increased sleep, a state termed “sickness sleep,” to promote recovery. The diagrams depict the gut lining with EECs highlighted, the interface with the BBB, and the resulting systemic effects on the organism’s sleep-wake states. EC; enterocyte. Statistical tests used: Unpaired two-sided t-tests for panel b, c and f; Ordinary ANOVA with Dunnett’s multiple comparisons for panel e, h, j, k, and l. Interaction effects were assessed using two-way ANOVA where indicated. Data are presented as mean ± SEM. ns, non-significant (p>0.05).

Next, we investigated whether AstA receptors are involved in mediating the glial-regulated sleep response to intestinal oxidative stress. Like the Unpaired cytokines, AstA is released from EECs49 and may therefore act as a context-dependent wake-promoting signal that, under certain conditions, is inhibited by Unpaired signaling in BBB glia to promote sleep. We thus hypothesized that during intestinal disturbances characterized by oxidative stress, gut-derived Unpaired signaling via JAK-STAT activation either sustains or consolidates sleep through a mechanism that involves the downregulation of wake-promoting AstA receptor signaling in BBB glial cells. In this model, EEC-derived Unpaired signaling normally suppresses AstA signaling in BBB glial cells under oxidative stress. Consequently, knocking down dome in these cells or unpaired in the EECs leads to a failure to downregulate AstA receptors, causing the animals to wake up under these conditions. Thus, inhibition of glial AstA receptors would impair the animals’ ability to respond to these wake-promoting signals altogether, leaving them unresponsive to intestinal ROS in terms of sleep. Consistent with this notion, we found that knocking down AstA-R1 or AstA-R2 in BBB glia attenuated the ROS-induced sleep response (Fig. S6b,c). Knockdown of AstA-R2 with second, independent RNAi line resulted in a more pronounced phenotype, with an almost completely blunted sleep response to intestinal ROS, showing no significant sleep increase during oxidative stress or the following day of recovery (Fig. 7d,e). AstA-R2 was also the most highly upregulated in response to loss of upd3 in the EEC or dome in glia (Fig. 7b,c), and it was the more strongly downregulated of the two AstA receptors in response to glial JAK-STAT activation, together suggesting that AstA-R2 is a primary receptor mediating these effects. Furthermore, knockdown of AstA receptors in BBB glia increased daytime sleep under normal homeostatic conditions, consistent with a wake-promoting role of AstA signaling in BBB glial (Fig. 7d,e and S6b,c).

AstA is produced by two cell types, neurons and the EECs in the gut49. Since BBB glial cells are well-positioned to receive hormonal signals from the periphery, they likely are regulated by gut-derived AstA. We thus examined whether gut-derived AstA acts as a wake-promoting signal, by conducting AstA knockdown in AstA-positive EECs using an AstA:2A::GAL4 knock-in in combination with R57C10-GAL80 (AstAGut>) to suppress GAL4 activity in the AstA-positive neuronal population. We confirmed that this driver efficiently reduces the expression of AstA in midguts without affecting neuronal AstA transcripts levels (Fig. 7f). Knockdown of AstA in AstA-positive EECs with either of two independent RNAi constructs led to increased sleep without any contribution of the transgenic insertion backgrounds (Fig. 7g,h and S6d,e), indicating that gut-derived AstA is indeed a wake-promoting factor. To assess whether EEC-derived AstA is sufficient to promote arousal, we employed the thermosensitive cation channel Transient Receptor Potential A1 (TrpA1)51 to induce hormonal release from AstA-positive EECs. Activation of these EECs suppressed sleep, an effect that was abolished by simultaneous AstA knockdown, supporting that the wake-promoting role of EEC-derived AstA (Fig. 7i,j).

Taken together, our findings suggest that enteric oxidative stress induces the release of Unpaired cytokines from the endocrine cells of the gut, which activate the JAK-STAT pathway in subperineurial glia of the BBB surrounding the brain. This activation leads to the glial downregulation of receptors for AstA, which is a wake-promoting factor also released by EECs. Gut-derived Upd signaling thereby gates the effect of AstA at the BBB and permits increased sleep during periods of intestinal stress. We therefore next investigated whether oxidative stress might also regulate the release of AstA from EECs. Following oxidative stress (24 hours of H2O2 feeding and the subsequent day), when wild-type animals exhibit increased sleep (Fig. 7e), AstA transcript levels in the midgut were reduced, accompanied by an accumulation of AstA peptide (Fig. 7k-m). This pattern – increased AstC staining in source cells despite decreased expression – suggests that oxidative stress suppresses AstA expression and release. This observation is consistent with a model in which, under conditions of enteric oxidative stress, wake-promoting gut-to-brain AstA signaling is silenced both at the source (gut EECs) and at the target (the BBB glia) by ROS-induced Unpaired signaling. We recently showed that Tk-positive EECs, which make up a population distinct from the AstA-positive EECs, express TrpA1, a ROS-sensitive cation channel known to promote hormone release, and thus exhibit ROS-induced Tk release24. Contrasting with that system, single-cell RNA sequencing data52 show that the AstA-expressing EECs do not express TrpA1 (Fig. S6f). This absence is consistent with our observation that oxidative stress does not promote AstA release (and indeed appears to inhibit it through mechanisms that remain to be explored), reinforcing the idea that gut-derived AstA signaling is actively suppressed rather than stimulated under these conditions. Together, these data support a model in which oxidative stress downregulates wake-promoting AstA signaling in the gut and simultaneously induces Unpaired cytokine signaling, which acts on BBB glia to suppress AstA receptor expression and thus to block the further transduction of wakefulness-promoting AstA signals. This dual-site regulation likely serves to silence arousal signals and promote sleep as a protective response to intestinal stress. This process may aid in recovery and maintain overall organismal homeostasis.

Discussion

Intestinal inflammation and microbial imbalance are strongly associated with sleep disturbances and mental disorders such as anxiety and depression8,12-14. The influence of gut health on CNS-dependent behaviors is thought to be mediated by the gut-brain axis, comprised of diverse signals secreted by the gut that act on the brain to induce behavioral responses53. Whereas the regulation of feeding behavior by this axis has been extensively studied, leading to revolutionary approaches to medical weight loss and diabetes control, the role of gut-brain signaling in regulating sleep – a behavior affected by nearly all illnesses54,55– remains poorly defined. Sickness induces a state of sleepiness, which is believed to be a conserved adaptive response that promotes recovery by supporting energy conservation and efficient immune activity5,6. However, the exact mechanisms driving sickness-induced sleep remain largely elusive. We have demonstrated here that intestinal ROS stress, through driving the release of interleukin 6-like Unpaired cytokines from endocrine cells of the Drosophila gut, regulates sleep via a glia-mediated pathway. This gut-to-glia communication promotes sleep during intestinal insult, presumably to facilitate the restorative sleep essential for both physical and mental health. Our findings provide mechanistic insight into how perturbations of gut health can influence sleep, potentially contributing to understanding the link between gastrointestinal disorders, sleep disturbances, and mental illnesses.

Cytokines, key secreted mediators of immune and inflammatory responses, are thought to modulate sleep/wake cycles under disease conditions56. Interleukins and TNFα, cytokines induced during illness in mammals, have been suggested to promote sleep to aid recovery from disease. However, most of these effects have been attributed to the actions of cytokines produced within the CNS, leaving open the question of how diseases affecting other parts of the body can drive sleep responses. In Drosophila, sleep induced by immune responses is known to be influenced by the NFκB ortholog Relish in fat tissue57, and the neuronally expressed gene nemuri drives sleep and connects immune function with sleep regulation6. However, inter-organ signaling mechanisms by which intestinal disease or stress regulate sleep have not yet been described in either flies or mammals. Intestinal infection or inflammation leads to elevated levels of ROS in the gut, and our findings demonstrate that enteric oxidative stress in the gut triggers the production of Upd2 and Upd3 cytokines by hormone-secreting EECs. These gut-derived cytokines signal the state of the intestine to brain glial cells, including those of the BBB, and modulate sleep. This glia-mediated gut-to-brain signaling promotes wakefulness in healthy animals under normal conditions, while inducing sleep in response to oxidative stress in the intestine. This indicates a dual functionality, with low levels of gut Unpaired signaling promoting wakefulness and higher stress-induced levels acting to enhance sleep (Fig. 7n). A similar dose-dependent effect has previously been observed for interleukins in rats, in which injection of IL-1 into the CNS can either stimulate sleep or inhibit it, depending on the administered dose58. While our findings show that ROS-induced cytokine signaling in the gut modulates sleep through gut–brain communication, an intriguing direction for future research will be to determine whether pathogenic infections – which trigger both intestinal ROS and additional immune pathways – engage distinct, complementary, or overlapping mechanisms compared to chemically induced oxidative stress, and how these immune responses collectively influence sleep regulation.

Our results indicate that Unpaired signaling in subperineurial glial cells – those forming the BBB – activates the JAK-STAT pathway, and they suggest that this effect inhibits wake-promoting AstA signaling by downregulating AstA receptor expression. AstA and its receptors, which are orthologous with the mammalian Galanin signaling system, have been linked to the regulation of sleep, feeding, and metabolism49,59,60. Mammalian glia also express receptors for Galanin61, which also regulates sleep62,63, further underscoring a conserved role in sleep modulation across species. AstA-producing neurons induce sleep by releasing glutamate onto sleep-regulatory neuronal circuits, although recent findings also suggest a wake-promoting role for AstA signaling50. Irrespective of neuronal AstA, our experiments clearly show that AstA released from EECs of the gut acts as a wake-promoting signal and that activation of AstA receptor signaling in BBB glial cells induces wakefulness. This highlights the potential of peptide hormones to elicit different effects depending on their source tissue and thus their accessible target cells – whether they are produced by the gut outside the BBB or by the CNS inside the barrier. A similar phenomenon has been demonstrated for neuropeptide F39,64. Our findings further suggest that AstA release from EECs is downregulated under oxidative stress in the gut, indicating that this wake-promoting signal is suppressed both at the level of the intestine and at the BBB via Unpaired cytokine signaling. This coordinated downregulation may serve to effectively silence this arousal pathway and promote sleep during intestinal stress.

Interestingly, intestinal ROS can also be generated as a consequence of sleep deprivation65,66, suggesting a potential feedback mechanism. This raises the possibility that ROS produced during sleep loss engages the same Unpaired–JAK-STAT signaling cascade described here, leading to suppression of gut-derived AstA signaling and facilitating recovery sleep. This model provides a mechanistic link between sleep deprivation, intestinal stress, and the regulation of sleep, and suggests that ROS may serve as a physiological signal integrating peripheral stress and behavioral state.

While our study investigated the effects of ROS induction, contrasting findings have been reported under conditions of antioxidant treatment65. Our data show both decreased AstA transcript levels and increased AstA peptide accumulation following oxidative stress – a combination typically interpreted as reduced production coupled with peptide retention24,25,39. In contrast, the reported increase in AstA peptide levels under antioxidant treatment was not accompanied by expression data65, making it difficult to determine whether the AstA accumulation under these conditions reflects enhanced retention and/or increased production. Furthermore, single-cell RNA sequencing data52 indicate that AstA-positive EECs do not express the ROS-sensitive cation channel TrpA1, supporting our observation that intestinal ROS does not stimulate AstA release. We recently found that TrpA1 is expressed in a distinct population of Tk-positive EECs and drives ROS-dependent release of the gut hormone Tk from these cells in Drosophila. This mechanism was also observed in the mammalian intestine24. In contrast, a previous report suggested TrpA1-dependent AstA release from EECs65, highlighting a potential discrepancy in whether this channel regulates AstA secretion. These differences may reflect context-specific variationin enteroendocrine function, and in any case they underscore the complexity of AstA regulation under varying conditions of gut stress.

Drosophila exhibit conserved behaviors such as sleep, arousal/wakefulness, and anxiety-like responses4,67-72, and the EECs of the fly gut produce diverse hormones similar to those of mammals17-22, potentially influenced by diet, microbiota, and inflammatory responses. This makes Drosophila an excellent model for studying behaviors influenced by gut conditions through gut-brain signaling. Our findings suggest that the oxidative-stress level within gut tissues, which is modulated by intestinal bacteria and immune activity26,27, regulates sleep via EEC-derived Unpaired signaling, potentially explaining the observed links between gut microbiota and sleep disturbances in both flies and humans13,73. Furthermore, in mammals, conditions such as inflammatory bowel disease that are linked with oxidative stress74 are often associated with sleep and mental-health disturbances8,12,14. Our results imply that cytokines, including interleukin signaling from an inflamed or diseased gut, might be a mechanism by which intestinal illnesses affect sleep and mental health. Our findings raise the possibility that these cytokines may act on glial cells that integrate and relay these gut signals to brain sleep-regulatory circuits.

The neurons of the CNS are isolated from the circulatory system by the BBB75 that restricts the transmission of some hormonal and cytokine signals from the periphery to neurons within the brain. Our work suggests that the BBB receives AstA and Unpaired signaling from the periphery. Other reports indicate that Unpaired cytokines from tumors and from enterocytes also can activate JAK-STAT signaling in BBB glia cells in Drosophila33,76. Although our findings highlight endocrine EECs as a primary source of gut-derived cytokines that act on the brain to regulate sleep, it is also possible that enterocytes or other non-endocrine gut cell types contribute to the systemic Unpaired signaling that modulates sleep in response to intestinal oxidative stress. One effect of glial JAK-STAT activity seems to be the alteration of BBB permeability76, raising the possibility that EEC-derived Unpaired signaling in BBB glia, directly or through AstA signaling, modulates sleep via regulation of BBB permeability, which has been linked to homeostatic sleep regulation36. Furthermore, the endocytic activity of BBB glia, important for cellular transport and barrier function, has also been associated with sleep regulation37, and thus JAK-STAT-induced changes could regulate sleep through alterations in intracellular trafficking within the cells of the BBB. Another possibility is that JAK-STAT activity might regulate glial metabolic support for neuronal activity and in this way affect sleep patterns. In any case, our findings highlight the involvement of BBB glial cells in transmitting signals from the gut to the brain, adding another layer to our understanding of body-to-brain communication, which suggests that the BBB does more than protect the brain; it also responds to peripheral signals to modulate brain function, presenting an intriguing area for future research into gut-brain signaling.

Methods

Drosophila stocks and husbandry

Flies were cultured using a standard cornmeal-based formulation (82 g/L cornmeal, 60 g/L sucrose, 34 g/L yeast, 8 g/L agar, 4.8 mL/L propionic acid, and 1.6 g/L methyl-4-hydroxybenzoate) maintained at 25 °C with 60% relative humidity under a 12-hour light/dark cycle. Post-eclosion, flies were transitioned to an adult-specific, cornmeal-free diet (comprising 90 g/L sucrose, 80 g/L yeast, 10 g/L agar, 5 mL/L propionic acid, and 15 mL/L of a 10% methyl-4-hydroxybenzoate solution in ethanol)77 for 4-7 days prior to experiments. Adult mated females were used for all experiments. Flies were separated by sex one day prior to experimental procedures. Strains harboring the temperature-sensitive Tubulin-GAL80ts transgene were initially reared at 18 °C on cornmeal food and then switched to the adult diet for 3-4 days post-eclosion, still at 18 °C. Subsequently, they were incubated at 29 °C for five to seven days to activate RNAi expression in advance of the experiments. To ensure optimal conditions, the flies were provided with fresh food every three days. The following lines used in this study were sourced from the Bloomington Drosophila Stock Center (BDSC) at the University of Indiana: R57C10-GAL4 (#39171); UAS-upd3-RNAiTRiP (#32859); UAS-dome-RNAiTRiP (#53890); AstA::2A::GAL4 (#84593); AstA-R1::2A::GAL4 (#84709); AstA-R2::2A::GAL4 (#84594); UAS-AstA-RNAiTRiP (#25866); UAS-AstA-R2-RNAiTRiP (#67864); UAS-mCD8::GFP (#5137); Tub-GAL80ts (#7108); repo-GAL4 (#7415); moody-GAL4 (#90883); UAS-TrpA1 (#26263); UAS-dsRed was extracted from (#8546); AstC::2A::GAL4 (#84595); Tk::2A::GAL4 (#84693); and Upd3-GAL4, UAS-GFP (#98420; GFP variant and protein localization are unknown in this line). Additional fly lines were acquired from the Vienna Drosophila Resource Center (VDRC): control line w1118 (#60000, which is isogenic with the VDRC RNAi lines); UAS-upd2-RNAiSH (#330691); UAS-upd3-RNAiKK (#106869); UAS-upd3-RNAiGD(#27136); UAS-dome-RNAiKK (#106071); UAS-dome-RNAiGD (#36356); UAS-AstA-RNAiKK (#103215); UAS-AstA-R1-RNAiKK (#101395); UAS-AstA-R2-RNAiKK(#108648). The upd3Δ and upd2,3Δ deletion mutants were kindly provided by Bruno Lemaitre. UAS-upd3 and UAS-hopTum lines were gifts from David Bilder. The STAT-dGFP:2A::RFP line was generously supplied by Norbert Perrimon. The voilà-GAL4 strain was graciously provided by Alessandro Scopelliti. The R57C10-GAL80 transgene, situated on the X chromosome, was kindly donated by Ryusuke Niwa. The 10xSTAT::GFP line was a gift from Julien Colombani. To ensure uniformity in genetic background and to create control groups with an appropriate genetic background, all GAL4 and GAL80 lines used in this study were backcrossed to a w1118 line for multiple generations before being outcrossed with the genetic background specific to the RNAi, CRISPR, or overexpression lines to serve as controls in the experiments25. This ensures that the only difference between experimental and control animals is the presence or absence of the UAS transgene, providing the most appropriate control for assessing transgene-specific effects.

Generation of tissue-specific CRISPR lines

To facilitate tissue-specific CRISPR-based disruption of the upd2 and upd3 loci, constructs were prepared containing two gRNA target sequences, flanked by efficiency-enhancing tRNA sequences. One construct was prepared for upd2, and two transgenes, targeting different genomic sites, were made for upd3. The upd2 construct was designed to delete the region encoding the secreted Upd2 protein. One upd3 construct should delete the initiator ATG codon, and the other – the one used in this work – deletes the second exon, which contains a significant portion of the coding sequence. Target-sequence cassettes were assembled by first cloning the tRNA insert from plasmid pCFD6 (Addgene #73915) between long oligos containing the gRNA target sequences using Q5 polymerase (New England Biolabs, #M0491S). The vector and the PCR products were then integrated using Gibson assembly (NEBuilder HiFi DNA Assembly Master Mix, New England Biolabs, #E2621S). Clones were sequenced to verify accuracy, and correct constructs were integrated into the fly genome at the attP2 site (chromosome 3L) by BestGene (Chino Hills, CA). The sequences used for cloning the upd2 and upd3 CRISPR constructs, with gRNA sequences indicated in bold, are shown in Supplementary Table 1.

Sleep, activity, and survival assays

The Drosophila Activity Monitoring System (TriKinetics, Waltham, MA) was employed to track sleep and activity patterns. Single flies aged six to eight days after eclosion were placed into glass tubes using light CO2 anesthesia. On one end, the tubes were sealed with a foam plug; on the other was placed a detachable 250-μL PCR tube containing 90 μL of feeding medium: either 5% sucrose in 1% agar/water, 5% sucrose mixed with various concentrations of H2O2 in 1% agar/water, or plain 1% agar/water for starvation conditions. All food media contained 0.5% propionic acid and 0.15% methyl-4-hydroxybenzoate to prevent microbial growth, with H2O2 being supplemented once the food had cooled to below 40 °C. Monitoring of the flies’ locomotor activity and sleep began at the beginning of the light cycle, after the animals had spent their first day in the tubes acclimating. Following an additional 24 hours on the standard 5%-sucrose diet, the PCR tubes were replaced with fresh ones containing H2O2 or starvation media at the lights-on transition when animals were awake, to avoid unnecessary disturbances to the animals. For recovery experiments, animals were switched back to a 5%-sucrose diet after 24 hours on H2O2-laced food. Periods of inactivity lasting 5 minutes or longer were recorded as “sleep.” In the sleep deprivation studies, the flies were placed in DAM monitors and subjected to mechanical stimulation, which was produced by attaching the monitors to a vortexer mounting plate (TriKinetics) and vibrating them for 2 seconds at the start of each minute throughout the 6-hour interval leading up to the lights-on time. Recovery sleep was assessed in flies that experienced a reduction of more than 60% in their typical sleep during the deprivation period, using their sleep patterns from the 24-hour period before the onset of sleep deprivation as a baseline. The occurrence of recovery sleep was specifically evaluated during the first two hours immediately following the sleep deprivation phase. For survival assays, flies were loaded into tubes filled with either plain 1% agar/water for starvation or 1% H2O2 in 1% agar/water to test oxidative stress resistance. The time of death was recorded upon the complete cessation of movement.

Feeding Assays

Short-term food consumption was quantified using a spectrophotometric dye-feeding assay78,79. All food intake experiments were conducted during the time of the normal morning meal, one hour after lights-on in a 12:12-hour dark/light cycle. Flies were transferred without anesthesia to food (90 g/L sucrose, 80 g/L yeast, 10 g/L agar, 5 mL/L propionic acid, and 15 mL/L of a 10% methyl-4-hydroxybenzoate solution in ethanol) supplemented with 0.5% erioglaucine dye (brilliant blue R, FD&C Blue No.1, Sigma-Aldrich, #861146) and allowed to feed for 1 hour. A control group of flies was provided with undyed food to establish the baseline absorbance levels of fly lysates. For each genotype, 1-2 flies per sample were homogenized in 100 μL phosphate buffer (pH 7.5) using a TissueLyser LT (Qiagen) with 5-mm stainless-steel beads. Homogenates were centrifuged at 16,000 g for 5 minutes, and 50 μL of the cleared supernatant was transferred to a 384-well plate. Absorbance was measured at 629 nm for erioglaucine using an Ensight multi-mode plate reader (PerkinElmer). Standard curves for dye were employed to correlate absorbance readings with the amounts of food consumed.

To assess feeding behavior, interactions with food were monitored over a 20-24 hour period using the Fly Liquid-Food Interaction Counter (FLIC) apparatus80. Drosophila Feeding Monitors (DFMs; Sable Systems) were placed in an incubator set to 25 °C (or 29 °C for strains carrying GAL80ts), maintaining 70% humidity under a 12:12-hour light/dark cycle. Each of the 12 DFM chambers was filled with a 10% sucrose solution, and individual flies were introduced in the afternoon following the morning meal. After several hours of acclimation, evening feeding activity was recorded. The following morning, at lights-on, the DFMs were refilled with fresh sugar solution, and data from the morning meal were collected. The feeding behavior was recorded using the manufacturer’s software and analyzed using R Studio with the provided package (https://github.com/PletcherLab/FLIC_R_Code).

Immunohistochemistry, TUNEL staining, and confocal imaging

Adult midguts, brains, and VNCs were dissected in cold PBS and fixed for 1 hour at room temperature in 4% paraformaldehyde/PBS with gentle shaking. After a quick rinse with PBST (PBS with 0.1% Triton X-100, Merck #12298), the tissues were washed three times for 15 minutes each in PBST. For TUNEL staining, the In Situ Cell Death Detection Kit, Fluorescein (Roche, #11684795910), was used according to the manufacturer’s instructions, and tissues were subsequently washed in PBS before mounting. For samples undergoing antibody staining, tissues were then blocked for 30 minutes at room temperature in PBST containing 5% normal goat serum (Sigma) and subsequently incubated overnight (or two days for CNS samples) at 4 °C with primary antibodies diluted in the blocking solution with gentle agitation. After removing the primary antibody solution, tissues were rinsed once and washed three times for 20 minutes each in PBST. Secondary antibodies diluted in PBST were applied, and tissues were incubated overnight at 4 °C, followed by three PBST washes and one PBS wash. The samples were then mounted on poly-L-lysine-coated slides (Sigma, #P8920) in Fluoroshield mounting medium with DAPI (Nordic Biosite, #GTX30920), and imaged on a Zeiss LSM-900 confocal microscope using a 20× air or 40× oil objective with Zen software. Image stitching was performed using the Stitching function of Zeiss Zen Blue 3.1, and analysis was conducted using the open-source FIJI/ImageJ software package81. All samples compared with each other within a figure panel were dissected, stained, and imaged simultaneously using identical settings and reagents. For quantification of AstA peptide levels in the R5 region of the posterior midgut, anti-AstA stained images were processed in FIJI, and mean fluorescence intensity was measured in the R5 region. Background signal was subtracted using neighboring non-AstA-expressing areas from the same tissue. To quantify 6xSTAT-dGFP::2A::RFP expression (temporally resolved JAK/STAT activity indicator), dissected brains were fixed for 10 minutes in 4% paraformaldehyde at room temperature with agitation, rinsed once with PBST, mounted, and imaged immediately without antibody staining. For 10xSTAT-GFP quantification, brain and VNC samples were stained as described above, Z-stacks were projected in FIJI using the “sum” method, and Repo-positive cells at the surface were manually segmented to measure the raw integrated density with local background subtraction. For measuring 10xSTAT-GFP across the BBB, linear regions of interest were drawn through the glial layer perpendicular to the brain surface at the plane showing maximum brain size using FIJI’s line tool; GFP intensity was then quantified along these lines, and the peak was recorded for each transect. Antibodies used included 1:2000 rabbit anti-AstA (Jena Bioscience, #ABD-062), 1:50 mouse anti-Repo (University of Iowa Developmental Studies Hybridoma Bank, #8D12), 1:20 mouse anti-Prospero (University of Iowa Developmental Studies Hybridoma Bank, #MR1A), 1:500 mouse anti-GFP (ThermoFisher, #A11120), 1:500 chicken anti-GFP (ThermoFisher, #A10262), 1:1000 rat anti-mCherry (used against Ds Red, ThermoFisher, #M11217), 1:500 Alexa Fluor 488-conjugated goat anti-mouse (ThermoFisher, #A11001), 1:500 Alexa Fluor 555-conjugated goat anti-mouse (ThermoFisher, #A21422), 1:500 Alexa Fluor 647-conjugated goat anti-mouse (ThermoFisher, #A32728), 555-conjugated goat anti-rat (ThermoFisher, #A21434), and 488-conjugated goat anti-chicken (ThermoFisher, #A11039).

In-situ hybridization

To detect co-expression of specific mRNAs, we performed fluorescent in situ hybridization using the hybridization chain reaction (HCR) method. This approach, based on a previously published protocol82 with slight modifications, utilizes fluorescent probes and reagents from Molecular Instruments (Los Angeles, CA, USA). Adult Drosophila midguts were dissected in ice-cold PBS and fixed in 4% paraformaldehyde (PFA) for 1 hour at room temperature with gentle rocking. After fixation, tissues were rinsed and washed three times for 10 minutes each at room temperature in PBS containing 0.1% Tween-20 (Sigma, #P1379). To permeabilize the tissue, samples were incubated at 37 °C for 30 minutes in a buffer containing 1% SDS, 0.5% Tween-20, 50 mM Tris-HCl, 1 mM EDTA, and 150 mM NaCl at pH 7.5. Tissues were then incubated in pre-warmed hybridization buffer (from the HCR buffer kit) for 0 minutes at 37 °C. Hybridization was carried out overnight at the same temperature using hybridization buffer containing 15 nM each of B3-labeled upd3 probes and B5-labeled prospero probes (Molecular Instruments). The next day, tissues were washed four times for 15 minutes each with pre-warmed probe wash buffer, followed by two 5-minute washes in 5× SSCT (prepared using 20× SSC concentrate [Sigma, #S6639] with 0.1% Tween-20 [Sigma, #P1379]). Amplification was initiated by incubating tissues in amplification buffer for 40 minutes at room temperature, followed by an overnight incubation with 120 nM of each corresponding fluorophore-labeled hairpin (Alexa Fluor 488-B3 and Alexa Fluor 546-B5), in the dark at room temperature. After amplification, tissues were washed five times for 10 minutes each in 5× SSCT, followed by six 15-minute washes in PBS. Midguts were mounted on poly-L-lysine–coated slides (Sigma, #P8920) using ProLong Glass Antifade Mountant (Invitrogen, #36984), and coverslipped using a 0.12-mm spacer and 0.1-mm glass coverslip. Imaging was performed with a Zeiss LSM-900 confocal microscope using Zen software.

Triacylglyceride measurements

Triacylglyceride concentrations were determined following established methods77,83 using the Randox Triglycerides (GPO-PAP) method (Randox, #TR210). For each sample, flies were homogenized in 50 μl PBS per fly (between 2-4 flies per sample) containing 0.1% Tween-20 (Sigma #1379) using a TissueLyser LT (Qiagen) with 5-mm stainless-steel beads, 50 oscillations/second for 30 seconds. Homogenates were heated at 70 °C for 10 minutes to inactivate endogenous enzymes and centrifuged at 11,000 g for 1 min. Aliquots of cleared and vortexed supernatants (4 μl) were added to 36 μl of triglyceride reagent (Randox, #TR210) in a 384-well plate, covered with ultra-high-clarity optical film (ThermalSeal RT2RRTM, Z722553, Excel Scientific). The plate was spun down at 1,500 g for 1 min to settle fluids and eliminate bubbles and incubated for 10 minutes at room temperature. Absorbance for each sample was measured at 540 nm on an Ensight multimode plate reader (PerkinElmer). The readings were then converted to TAG concentrations using standard curves, prepared with triglyceride standards (Randox, 1352TR CAL Standard).

Measurement of transcript levels using qPCR

Several tissue samples containing 3 dissected guts, brains, or heads were collected for each condition or genotype. These samples were then homogenized in 2-ml Eppendorf tubes filled with lysis buffer containing 1% beta-mercaptoethanol, utilizing a TissueLyser LT bead mill (Qiagen) with 5-mm stainless steel beads (Qiagen #69989). RNA extraction was carried out with the NucleoSpin RNA kit (Macherey-Nagel, #740955) following the guidelines provided by the manufacturer. cDNA was synthesized using the High-Capacity cDNA Synthesis kit (Applied Biosystems, #4368814). Quantitative PCR was performed with RealQ Plus 2x Master Mix Green (Ampliqon, #A324402) using a QuantStudio 5 (Applied Biosystems) instrument. Gene expression results were normalized to the housekeeping gene Rp49 using the delta-delta-Ct method. The specific oligonucleotides used are given in Supplementary Table 2.

Statistics

Statistical analyses were performed using the Prism software package (GraphPad, version 10). Data were tested for normality before assessments of significance. For data following a normal distribution, pairwise analyses were conducted using two-tailed unpaired Student’s t-tests, and comparisons involving multiple samples used one-way ANOVA with subsequent post-hoc tests for multiple comparisons. Non-normally distributed data were analyzed using two-tailed unpaired Mann-Whitney U tests or one-way Kruskal-Wallis ANOVA, followed by multiple comparisons. Additionally, interactions between genotype and diet were calculated using two-way ANOVA. All plots represent the mean ± standard error of the mean (SEM). All replicates represent independent biological samples.

Data Availability

All data supporting the findings of this study will be freely available from the lead contact upon request, without restriction.

Code Availability

The MatLAB scripts used for analyzing sleep are described10,84.